---
title: "Integrative multiomic profiling reveals epigenetic precoding, transcription factor hierarchies and reprogrammability of the senescent state"
subtitle: "Appendix C : Analyses of ATAC-seq data"
date: ""

output:
   BiocStyle::html_document:
    theme: cosmo
    highlight:  "tango"
    toc: true
    toc_float:
      toc_collapsed: true
    toc_depth: 2
    number_sections: true
    code_folding: 'show'
    df_print: paged
abstract: |
---
  
  \pagebreak

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE, cache.lazy = FALSE, eval = TRUE, results = "hide", message = FALSE, warning = FALSE, echo = TRUE, fig.wide = TRUE,  fig.align='center', cache.comments = FALSE)
```

# Overview of the data

The table below summarize the information relative to the histone modification ATAC-seq data anlyzed in the scope of this study. The information given are :
  
* *Time point* : T0 - before induction of RAS, or 24h-48h-72h-96h-144h after induction,
* *Replicate* : there are three replicates per time-point,
* *SRA* : Sequence Read Archive accession number,
* *Raw reads* : total number of reads for each library before any processing,
* *Clean reads* : number of reads remaining after quality filtering and adapter triming,
* *Uniquely mapped* : number of unambiguously mapped reads,
* *Non duplicated* : number of reads after filter for PCR / optical duplicates,
* *Not in BLR* : number of reads mapped outside hg19 ENCODE blacklisted regions (BLR),
* *% kept* : percentage of reads kept for subsequent analysis after filtering,
* *% duplicates* : percentage of PCR / optical duplicates,
* *% BLR* : percentage of reads mapped in BLR,
* *% chrM* : percentage of reads mapped on mitochondrial chromosome.

```{r, results= "asis", echo = FALSE}
library(dplyr)
library(knitr)
library(kableExtra)
info_atacseq <- read.table("./data/ATAC-seq/ATAC-seq_data_info.txt", header = TRUE, sep = "\t")

kable(info_atacseq,
      col.names = c("Time point", "Replicate", "SRA",
                    "Raw reads (N)", "Clean reads (N)", "Uniquely mapped (N)",
                    "Non duplicated (N)", "Not in BLR (N)", "% kept",
                    "% duplicates", "% in BLR", "% chrM")) %>%
  kable_styling("striped", full_width = TRUE, font_size = 9, position = "left") %>%
  collapse_rows(columns = 1:2, valign = "top")

```


# Preprocessing of ATAC-seq

All the ATAC-seq data were processed using a workflow written in *bash* and ran on a Unix cluster driven by Sun Gride Engine. The different steps we described bellow are :

* quality check on raw reads with [fastQC 0.11.5](https://www.bioinformatics.babraham.ac.uk/projects/fastqc/),
* read croping with [trimmomatic 0.33](http://www.usadellab.org/cms/?page=trimmomatic)
* read trimming and cleaning with [cutadapt 1.9.1](https://github.com/marcelm/cutadapt),
* check for contaminations with [fastq-screen 0.4.4](https://www.bioinformatics.babraham.ac.uk/projects/fastq_screen/),
* alignment with [bowtie2 2.2.3](http://bowtie-bio.sourceforge.net/bowtie2/index.shtml),
* de-duplication with [PicardTools 1.130](https://broadinstitute.github.io/picard/),
* filtering of reads aligned in [Encode blaclisted regions](https://sites.google.com/site/anshulkundaje/projects/blacklists) with [bedtools 2.19.1](http://bedtools.readthedocs.io/en/latest/),
* sorting and filtering of alignments with [samtools 1.6](http://samtools.sourceforge.net),
* peak calling with [MACS 2.1.0](https://github.com/taoliu/MACS),
* irreproducible discovery rate computation with [idr](https://sites.google.com/site/anshulkundaje/projects/idr).

Briefly, in this workflow, paired-ends reads were cropped to 100bp with trimmomatic v0.36 and cleaned using cutadapt v1.8.3 to remove Nextera adapters, low quality bases and reads, and discard reads shorter than 25 bp after filtering. Fragments were then aligned to the human reference genome (hg19) using bowtie2 v2.2.3 discarding inconsistent pairs and considering a maximum insert size of 2kb (bowtie2 -N 0 --no-mixed --no-discordant --minins 30 --maxins 2000). Alignment files were further processed with samtools v1.2 and PicardTools v1.130 to flag PCR and optical duplicates and remove alignments located in Encode blacklisted regions. Accessible regions were identified using MACS2 v2.1.0 without control (macs2 callpeak --gsize hs -p 1e-3). These relaxed peak lists were then processed through the irreproducible discovery rate (IDR) pipeline to generate an optimal and reproducible set of peaks for each time point.

## Preparing the working environment 

```{bash,  eval = FALSE}
##--> Create main directories
mkdir ./Data/
mkdir ./Output/
mkdir ./Scripts/
mkdir ./Std_Output/

##--> Create subdirectories
mkdir -p ./Data/0-Raw-Fastq/
mkdir -p ./Data/0-TRIMMED_Fastq/
mkdir -p ./Data/1-DEDUP_Bam/
mkdir -p ./Data/1-MAPPING_Bam/
mkdir -p ./Data/1-NOBLACKLIST_Bam/

mkdir -p ./Output/0-CONTAMINATION/
mkdir -p ./Output/0-FAST-QC/
mkdir -p ./Output/3-NucleoATAC/

mkdir -p ./Std_Output/0-FastqCUT/
mkdir -p ./Std_Output/0-FastqMCF/
mkdir -p ./Std_Output/0-FastqScreen/
mkdir -p ./Std_Output/1-Bowtie/
mkdir -p ./Std_Output/1-Sort/
mkdir -p ./Std_Output/1-Markdup/
mkdir -p ./Std_Output/1-Dedup/
mkdir -p ./Std_Output/1-NoBL_Dedup/
mkdir -p ./Std_Output/3-MACS2/
```

## QC on row data 

```{bash, eval = FALSE}
##--> Run fastQC on all samples 
for i in $(ls --color=never ./Data/0-Raw-Fastq/)
do
    qsub -q ono -cwd -V -l mem_rsvd=8G -b y fastqc --nogroup ./Data/0-Raw-Fastq/$i
done

##--> Move outputs
mv ./Data/0-Raw-Fastq/*fastqc* ./Output/0-FAST-QC/

```

## Apdapters cleaning, read triming and filtering

```{bash, eval = FALSE}
##--> Creation of batch script : crop to 100 bp
for i in $(ls --color=never ./Data/0-Raw-Fastq/ | grep "fastq.gz$" | sed 's/_S[12]//g' |sed 's/.fastq.gz//g' | sort -u)
do
    echo "java -jar trimmomatic-0.35.jar \
            PE \
            ./Data/0-Raw-Fastq/"$i"_S1.fastq.gz \
            ./Data/0-Raw-Fastq/"$i"_S2.fastq.gz \
            ./Data/0-TRIMMED_Fastq/"$i"_S1_TRIMMED.fastq.gz \
            ./Data/0-TRIMMED_Fastq/"$i"_S1_TRIMMED.fastq.gz.trash \
            ./Data/0-TRIMMED_Fastq/"$i"_S2_TRIMMED.fastq.gz \
            ./Data/0-TRIMMED_Fastq/"$i"_S2_TRIMMED.fastq.gz.trash \
            CROP:100" \
            >> ./Scripts/FASTQ-TRIMMOMATIC.txt
done

##--> Run the analysis
qarray \
    -q ono \
    -cwd \
    -V \
    -N FastqTRIMMO \
    -e ./Std_Output/0-FastqCUT/ \
    -o ./Std_Output/0-FastqCUT/ \
    ./Scripts/FASTQ-TRIMMOMATIC.txt
            

##--> Creation of batch script : Triming / Cleaning
for i in $(ls --color=never ./Data/0-Raw-Fastq/ | grep "fastq.gz$" | sed 's/_S[12]//g' |sed 's/.fastq.gz//g' | sort -u)
do
    echo "cutadapt \
            --overlap 10 \
            -q 30,30 \
            -n 3 \
            -a NNCTGTCTCTTATACACATCT \
            -A NNCTGTCTCTTATACACATCT \
            -g AGATGTGTATAAGAGACAG \
            -g CCGGCACATCTCCGAGCCCACGAGACNNNNNNNNATCTCGTATGCCGTCTTCTGCTTG \
            -G CCGGCACATCTCCGAGCCCACGAGACNNNNNNNNATCTCGTATGCCGTCTTCTGCTTG \
            -G AGATGTGTATAAGAGACAG \
            -m 25 \
            -o ./Data/0-TRIMMED_Fastq/"$i"_S1_TRIMMED_1.fastq.gz \
            -p ./Data/0-TRIMMED_Fastq/"$i"_S2_TRIMMED_1.fastq.gz \
            ./Data/0-TRIMMED_Fastq/"$i"_S1_TRIMMED.fastq.gz \
            ./Data/0-TRIMMED_Fastq/"$i"_S2_TRIMMED.fastq.gz" \
    >> ./Scripts/FASTQ-CUTADAPT.txt
done

##--> Run the analysis
qarray \
    -q ono \
    -cwd \
    -V \
    -N FastqCUT \
    -e ./Std_Output/0-FastqCUT/ \
    -o ./Std_Output/0-FastqCUT/ \
    ./Scripts/FASTQ-CUTADAPT.txt

##--> Creation of batch script :
## Remove 6 extra bp to avoid Bowtie2 issue with doved read (biased fragment length distribution)

for i in $(ls --color=never ./Data/0-Raw-Fastq/ | grep "fastq.gz$" | sed 's/_S[12]//g' |sed 's/.fastq.gz//g' | sort -u)
do
    echo "cutadapt \
            -u -20 -U -20 -m 25 \
            -o ./Data/0-TRIMMED_Fastq/"$i"_S1_TRIMMED_2.fastq.gz \
            -p ./Data/0-TRIMMED_Fastq/"$i"_S2_TRIMMED_2.fastq.gz \
            ./Data/0-TRIMMED_Fastq/"$i"_S1_TRIMMED_1.fastq.gz \
            ./Data/0-TRIMMED_Fastq/"$i"_S2_TRIMMED_1.fastq.gz" \
    >> ./Scripts/FASTQ-CUTADAPT_2.txt
done

##--> Run the analysis
qarray \
    -q ono \
    -cwd \
    -V \
    -N FastqCUT_2 \
    -e ./Std_Output/0-FastqCUT/ \
    -o ./Std_Output/0-FastqCUT/ \
    ./Scripts/FASTQ-CUTADAPT_2.txt

for i in $(ls --color=never ./Data/0-Raw-Fastq/ | grep "fastq.gz$" | sed 's/_S[12]//g' |sed 's/.fastq.gz//g' | sort -u)
do
    mv ./Data/0-TRIMMED_Fastq/"$i"_S1_TRIMMED_2.fastq.gz ./Data/0-TRIMMED_Fastq/"$i"_S1_TRIMMED.fastq.gz
    mv ./Data/0-TRIMMED_Fastq/"$i"_S2_TRIMMED_2.fastq.gz ./Data/0-TRIMMED_Fastq/"$i"_S2_TRIMMED.fastq.gz
    rm ./Data/0-TRIMMED_Fastq/"$i"_S1_TRIMMED_1.fastq.gz
    rm ./Data/0-TRIMMED_Fastq/"$i"_S2_TRIMMED_1.fastq.gz
done

##--> Run fastqc on Trimmed reads
for i in $(ls --color=never ./Data/0-TRIMMED_Fastq/)
do
    qsub -q ono -cwd -V -l mem_rsvd=8G -b y fastqc  --nogroup ./Data/0-TRIMMED_Fastq/$i
done

mv ./Data/0-TRIMMED_Fastq/*fastqc* ./Output/0-FAST-QC/

```

## Check for contamination

```{bash, eval = FALSE}
##--> Creation of batch script
for i in $(ls --color=never ./Data/0-Raw-Fastq/ | grep "fastq.gz$" | sed 's/\.fastq.gz//g')
do
    rm ./Output/0-CONTAMINATION/"$i"_TRIMMED_screen.txt
    echo fastq_screen                                          `# Version: 0.4.4` \
        --subset 300000                                        `# Number of reads for subsampling` \
        --outdir ./Output/0-CONTAMINATION/                     `# Output directory` \
        --aligner bowtie2                                      `# Aligner used (bowtie or bowtie2)` \
        --threads 4                                            `# Number of threads` \
       ./Data/0-TRIMMED_Fastq/"$i"_TRIMMED.fastq.gz \
    >> ./Scripts/FASTQ-SCREEN.txt
done

##--> Run the analysis
qarray \
    -q ono \
    -pe thread 4 \
    -cwd \
    -V \
    -N FastqScreen \
    -e ./Std_Output/0-FastqScreen/ \
    -o ./Std_Output/0-FastqScreen/ \
    ./Scripts/FASTQ-SCREEN.txt
```

## Alignment

```{bash, eval = FALSE}

# Creation of batch script
for i in $(ls --color=never ./Data/0-Raw-Fastq/ | grep "fastq.gz$" | sed 's/_S[12]//g' |sed 's/.fastq.gz//g' | sort -u)
do
    echo bowtie2                                                    `# Version: 2.2.3`\
        -p 8                                                         `# Number of threads`\
        -N 0                                                         `# Number of mismatches in the seed`\
        --no-mixed                                                   `# Only consider pairs`\
        --no-discordant                                              `# Don't consider discordant pairs`\
        --maxins 2000                                                `# Max insert size`\
        -x /local/databases/index/bowtie/hg19/2.1.0/hg19.fa          `# Reference index`\
        -1 ./Data/0-TRIMMED_Fastq/"$i"_S1_TRIMMED.fastq              `# Input S1`\
        -2 ./Data/0-TRIMMED_Fastq/"$i"_S2_TRIMMED.fastq              `# Input S2`\
        --trim3 8 \
    | samtools view -Sb - > ./Data/1-MAPPING_Bam/"$i".bam           `# Conversion sam to bam with samtools`\
>> ./Scripts/BOWTIE.txt
done

# Run the analysis
qarray \s
    -q ono \
    -cwd \
    -V \
    -pe thread 8 \
    -N BOWTIE \
    -e ./Std_Output/1-Bowtie/ \
    -o ./Std_Output/1-Bowtie/ \
    ./Scripts/BOWTIE.txt
```

## Sorting

```{bash, eval = FALSE}
##--> Creation of batch script
for i in $(ls --color=never ./Data/0-TRIMMED_Fastq/ | grep "fastq$" | sed 's/_S[12]_TRIMMED_2.fastq//g' | sort -u)
do
    echo samtools sort -@ 8 ./Data/1-MAPPING_Bam/"$i".bam ./Data/1-MAPPING_Bam/"$i"_SORT \
    >> ./Scripts/SAMTOOLS_SORT.txt
done

##--> Run the analysis        
qarray \
    -q ono \
    -cwd \
    -V \
    -pe thread 4 \
    -N Sort \
    -e ./Std_Output/1-Sort/ \
    -o ./Std_Output/1-Sort/ \
    ./Scripts/SAMTOOLS_SORT.txt

```

## De-duplication

```{bash, eval = FALSE}
# Creation of batch script
for i in $(ls --color=never ./Data/0-Raw-Fastq/ | grep "fastq.gz$" | sed 's/\.fastq.gz//g'|sed 's/\_S[12]//g' | sort -u)
do
    echo java -jar -Xmx16g ~/bin/Tools/PicardTools_v1.130/picard.jar \
        MarkDuplicates \
        REMOVE_DUPLICATES=FALSE \
        METRICS_FILE=./Data/1-DEDUP_Bam/"$i"_MARKDUP.txt \
        INPUT=./Data/1-MAPPING_Bam/"$i"_SORT.bam \
        OUTPUT=./Data/1-DEDUP_Bam/"$i"_MARKDUP.bam \
    >> ./Scripts/MARKDUP.txt
done

# Run the analysis        
qarray \
    -q ono \
    -cwd \
    -l mem_rsvd=16G \
    -V \
    -N Markdup \
    -e ./Std_Output/1-Markdup/ \
    -o ./Std_Output/1-Markdup/ \
    ./Scripts/MARKDUP.txt

```

## Blacklisted regions

```{bash, eval = FALSE}
##--> Creation of batch script
for i in $(ls --color=never ./Data/1-DEDUP_Bam/ | grep "bam$" | sed 's/\_MARKDUP.bam$//g'| sort -u |grep Rep[123])
do
    echo " samtools sort -@ 8 -n -o ./Data/1-DEDUP_Bam/"$i"_MARKDUP.bam ./Data/1-DEDUP_Bam/"$i"_MARKDUP.tmp | \
           pairToBed \
                -type neither \
                -abam stdin \
                -b ~/save/Outils/consensusBlacklist_hg19.bed \
            > ./Data/1-NOBLACKLIST_Bam/"$i"_DEDUP_NOBLACKLIST.bam" \
     >> ./Scripts/NOBLACKLIST_DEDUP.txt
done


##--> Run the analysis
qarray \
    -pe thread 8 \
    -q ono \
    -cwd \
    -V \
    -N NoBL_Dedup \
    -e ./Std_Output/1-NoBL_Dedup/ \
    -o ./Std_Output/1-NoBL_Dedup/ \
    ./Scripts/NOBLACKLIST_DEDUP.txt
```

## Final sorting and indexation

```{bash, eval = FALSE}
##--> Creation of batch script
for i in $(ls --color=never ./Data/1-DEDUP_Bam/ | grep "bam$" | sed 's/\_MARKDUP.bam$//g'| sort -u |grep Rep[123])
do
    echo "samtools view -h -F 1548 -q 30 ./Data/1-NOBLACKLIST_Bam/"$i"_DEDUP_NOBLACKLIST.bam \
    | samtools view -hSb - \
    | samtools sort -@ 8 - ./Data/1-NOBLACKLIST_Bam/"$i"_DEDUP_NOBLACKLIST_SORT" \
    >> ./Scripts/SAMTOOLS_SORT_2.txt
done

##--> Run the analysis        
qarray \
    -q ono \
    -cwd \
    -V \
    -pe thread 8 \
    -N Sort \
    -e ./Std_Output/1-Sort/ \
    -o ./Std_Output/1-Sort/ \
    ./Scripts/SAMTOOLS_SORT_2.txt


for i in $(ls --color=never ./Data/0-Raw-Fastq/ | grep "fastq.gz$" | sed 's/\.fastq.gz//g'|sed 's/\_S[12]//g' | sort -u)
do
    qsub -q ono -cwd -V -b y samtools index ./Data/1-NOBLACKLIST_Bam/"$i"_DEDUP_NOBLACKLIST_SORT.bam
done
```    

## Peak calling

### Parameters

```{bash, eval = FALSE}
marks=(ATAC)
days=(D0 D1 D2 D3 D4 D6)
reps=(Rep2 Rep3)
genomeFile=/local/databases/index/samtools/hg19/0.1.19/hg19.fa
pvalPeaks=1e-3
```


### Peak calling on original replicates

```{bash, eval = FALSE}
##--> Creation of batch script 
for i in $(ls --color=never ./Data/0-Raw-Fastq/ | grep "fastq.gz$" | sed 's/_S[12]\.fastq.gz//g' |sort -u)
do
    echo "macs2 callpeak callpeak \
        --format BAMPE \
        --gsize hs \
        -n ./Data/3-PEAK-CALLING/MACS/FRAG_100/SHARP/"$i"_MACS2_FRAG_100_SHARP \
        --gsize hs \
        -p 1e-3 \
        --tempdir ./ -t ./Data/1-NOBLACKLIST_Bam/"$i"_DEDUP_NOBLACKLIST_SORT.bam" \
    >> ./Scripts/MACS2_FRAG_100_SHARP.txt
done

##--> Run the analysis
qarray \
    -q ono \
    -cwd \
    -V \
    -N MACS2 \
    -l mem_rsvd=16G \
    -e ./Std_Output/3-MACS2/ \
    -o ./Std_Output/3-MACS2/ \
    ./Scripts/MACS2_FRAG_100_SHARP.txt
```

### Peaks calling on pseudoreplicates of individual replicates

```{bash, eval = FALSE}
##--> STEP 1.1 : Count the number of alignments in the each replicated .bam and split them in 2 .bam

# Creation of batch script
for mark in ${marks[@]} ; do
 for day in ${days[@]} ; do
  for rep in ${reps[@]} ; do
     
     echo "\
     samtools view -h -F 4 -b ./Data/1-NOBLACKLIST_Bam/"${day}${mark}_${rep}"_DEDUP_NOBLACKLIST_SORT.bam > ./Data/3-IDR/"${day}${mark}_${rep}"_DEDUP_NOBLACKLIST_ONLYMAPPED.bam ;
     samtools index ./Data/3-IDR/"${day}${mark}_${rep}"_DEDUP_NOBLACKLIST_ONLYMAPPED.bam ;
     
     nlines=\$(samtools idxstats ./Data/3-IDR/"${day}${mark}_${rep}"_DEDUP_NOBLACKLIST_ONLYMAPPED.bam | awk '{s+=\$3} END {print s}') ; nlines=\$(( (nlines + 1) / 2 )) ;
     
     samtools view ./Data/3-IDR/"${day}${mark}_${rep}"_DEDUP_NOBLACKLIST_ONLYMAPPED.bam > ./Data/3-IDR/"${day}${mark}_${rep}"_DEDUP_NOBLACKLIST_ONLYMAPPED.sam ;
     ~/bin/Tools/sample/sample ./Data/3-IDR/"${day}${mark}_${rep}"_DEDUP_NOBLACKLIST_ONLYMAPPED.sam --shuffle | split -d -l \${nlines} - ./Data/3-IDR/"${day}${mark}_${rep}"_DEDUP_NOBLACKLIST_ONLYMAPPED ;" \
     >> ./Scripts/IDR_STEP1.1_${day}${mark}_${rep}.txt
     
  done ;
 done ;
done
 

# Run the analysis
for mark in ${marks[@]} ; do
 for day in ${days[@]} ; do
  for rep in ${reps[@]} ; do
    qsub \
        -q ono \
        -cwd \
        -V \
        -l mem_rsvd=32G \
        -N IDR_Step_1.1 \
        -e ./Std_Output/3-IDR/ \
        -o ./Std_Output/3-IDR/ \
        ./Scripts/IDR_STEP1.1_${day}${mark}_${rep}.txt
  done ;
 done ;
done
   
##--> STEP 1.2 :  Sort and index the 2 PR .bam 

# Creation of batch script
for mark in ${marks[@]} ; do
 for day in ${days[@]} ; do
  for rep in ${reps[@]} ; do
     
     echo "\
     samtools view -bT $genomeFile ./Data/3-IDR/"${day}${mark}_${rep}"_DEDUP_NOBLACKLIST_ONLYMAPPED00 > ./Data/3-IDR/"${day}${mark}_${rep}"_PR1_DEDUP_NOBLACKLIST_ONLYMAPPED.bam ;
     samtools sort -@ 4 ./Data/3-IDR/"${day}${mark}_${rep}"_PR1_DEDUP_NOBLACKLIST_ONLYMAPPED.bam ./Data/3-IDR/"${day}${mark}_${rep}"_tmp1 ;
     mv ./Data/3-IDR/"${day}${mark}_${rep}"_tmp1.bam ./Data/3-IDR/"${day}${mark}_${rep}"_PR1_DEDUP_NOBLACKLIST_ONLYMAPPED.bam ;
     samtools index ./Data/3-IDR/"${day}${mark}_${rep}"_PR1_DEDUP_NOBLACKLIST_ONLYMAPPED.bam ;
     #rm ./Data/3-IDR/"${day}${mark}_${rep}"_DEDUP_NOBLACKLIST_ONLYMAPPED00
     
     samtools view -bT $genomeFile ./Data/3-IDR/"${day}${mark}_${rep}"_DEDUP_NOBLACKLIST_ONLYMAPPED01 > ./Data/3-IDR/"${day}${mark}_${rep}"_PR2_DEDUP_NOBLACKLIST_ONLYMAPPED.bam ;
     samtools sort -@ 4 ./Data/3-IDR/"${day}${mark}_${rep}"_PR2_DEDUP_NOBLACKLIST_ONLYMAPPED.bam ./Data/3-IDR/"${day}${mark}_${rep}"_tmp2 ;
     mv ./Data/3-IDR/"${day}${mark}_${rep}"_tmp2.bam ./Data/3-IDR/"${day}${mark}_${rep}"_PR2_DEDUP_NOBLACKLIST_ONLYMAPPED.bam ;
     samtools index ./Data/3-IDR/"${day}${mark}_${rep}"_PR2_DEDUP_NOBLACKLIST_ONLYMAPPED.bam ;
     #rm ./Data/3-IDR/"${day}${mark}_${rep}"_DEDUP_NOBLACKLIST_ONLYMAPPED01" \
     >> ./Scripts/IDR_STEP1.2_${day}${mark}_${rep}.txt
     
  done ;
 done ;
done  

# Run the analysis
for mark in ${marks[@]} ; do
 for day in ${days[@]} ; do
  for rep in ${reps[@]} ; do
    qsub \
        -pe thread 4 \
        -q ono \
        -cwd \
        -V \
        -N IDR_Step_1.2 \
        -e ./Std_Output/3-IDR/ \
        -o ./Std_Output/3-IDR/ \
        ./Scripts/IDR_STEP1.2_${day}${mark}_${rep}.txt
  done ;
 done ;
done

##--> STEP 1.3 :  Peak calling on the two PR

# Creation of batch script
for mark in ${marks[@]} ; do
 for day in ${days[@]} ; do
  for rep in ${reps[@]} ; do
    for pr in PR1 PR2 ; do
        
        qsub \
            -l mem_rsvd=16G \
            -q ono \
            -cwd \
            -V \
            -N IDR_Step_1.3 \
            -e ./Std_Output/3-IDR/ \
            -o ./Std_Output/3-IDR/ \
            -b y macs2 callpeak callpeak --format BAMPE --gsize hs -p $pvalPeaks --tempdir ./ -t ./Data/3-IDR/${day}${mark}_${rep}_${pr}_DEDUP_NOBLACKLIST_ONLYMAPPED.bam -n ./Data/3-IDR/${day}${mark}_${rep}_${pr}_MACS2
    
    done ;
  done ;
 done ;
done

```

### Peak calling on pooled pseudoreplicates of induvidual replicates

```{bash, eval = FALSE}
##--> STEP 2.1 : Pool the two replicates and create two pooled PS

# Creation of batch script
for mark in ${marks[@]} ; do
 for day in ${days[@]} ; do
   
    echo "\
   #3.1 pool bam files IPs
   samtools view -H ./Data/3-IDR/"${day}${mark}"_Rep2_DEDUP_NOBLACKLIST_ONLYMAPPED.bam > ./Data/3-IDR/"${day}${mark}"_Header.txt ;
   samtools merge -h ./Data/3-IDR/"${day}${mark}"_Header.txt ./Data/3-IDR/"${day}${mark}"_POOLED_DEDUP_NOBLACKLIST_ONLYMAPPED.bam ./Data/3-IDR/"${day}${mark}"_Rep*_DEDUP_NOBLACKLIST_ONLYMAPPED.bam ;
   samtools index ./Data/3-IDR/"${day}${mark}"_POOLED_DEDUP_NOBLACKLIST_ONLYMAPPED.bam ;

   #3.2 get pseudo-replicates
   #Number of reads in the tagAlign file
   nlines=\$(samtools idxstats ./Data/3-IDR/"${day}${mark}"_POOLED_DEDUP_NOBLACKLIST_ONLYMAPPED.bam | awk '{s+=\$3} END {print s}') ; nlines=\$(( (nlines + 1) / 2 )) ;
   
    samtools view ./Data/3-IDR/"${day}${mark}"_POOLED_DEDUP_NOBLACKLIST_ONLYMAPPED.bam > ./Data/3-IDR/"${day}${mark}"_POOLED_DEDUP_NOBLACKLIST_ONLYMAPPED.sam
   ~/bin/Tools/sample/sample ./Data/3-IDR/"${day}${mark}"_POOLED_DEDUP_NOBLACKLIST_ONLYMAPPED.sam --shuffle | split -d -l \${nlines} -  ./Data/3-IDR/"${day}${mark}"_POOLED_DEDUP_NOBLACKLIST_ONLYMAPPED ;
   
   #pseudorep1
   samtools view -bT $genomeFile ./Data/3-IDR/"${day}${mark}"_POOLED_DEDUP_NOBLACKLIST_ONLYMAPPED00 > ./Data/3-IDR/"${day}${mark}"_POOLED_REP1_DEDUP_NOBLACKLIST_ONLYMAPPED.bam ;
   samtools sort -@ 4 ./Data/3-IDR/"${day}${mark}"_POOLED_REP1_DEDUP_NOBLACKLIST_ONLYMAPPED.bam ./Data/1-NOBLACKLIST_Bam/"${day}${mark}"_tmp1 ;
   mv ./Data/1-NOBLACKLIST_Bam/"${day}${mark}"_tmp1.bam ./Data/3-IDR/"${day}${mark}"_POOLED_REP1_DEDUP_NOBLACKLIST_ONLYMAPPED.bam ;
   samtools index ./Data/3-IDR/"${day}${mark}"_POOLED_REP1_DEDUP_NOBLACKLIST_ONLYMAPPED.bam ;
   
   #pseudorep2
   samtools view -bT $genomeFile ./Data/3-IDR/"${day}${mark}"_POOLED_DEDUP_NOBLACKLIST_ONLYMAPPED01 > ./Data/3-IDR/"${day}${mark}"_POOLED_REP2_DEDUP_NOBLACKLIST_ONLYMAPPED.bam ;
   samtools sort -@ 4 ./Data/3-IDR/"${day}${mark}"_POOLED_REP2_DEDUP_NOBLACKLIST_ONLYMAPPED.bam ./Data/1-NOBLACKLIST_Bam/"${day}${mark}"_tmp2 ;
   mv ./Data/1-NOBLACKLIST_Bam/"${day}${mark}"_tmp2.bam ./Data/3-IDR/"${day}${mark}"_POOLED_REP2_DEDUP_NOBLACKLIST_ONLYMAPPED.bam ;
   samtools index ./Data/3-IDR/"${day}${mark}"_POOLED_REP2_DEDUP_NOBLACKLIST_ONLYMAPPED.bam" \
   >> ./Scripts/IDR_STEP2.1_${day}${mark}.txt

 done ;
done

# Run the analysis
for mark in ${marks[@]} ; do
 for day in ${days[@]} ; do
    qsub \
        -pe thread 4 \
        -l mem_rsvd=32G \
        -q ono \
        -cwd \
        -V \
        -N IDR_Step_2.1 \
        -e ./Std_Output/3-IDR/ \
        -o ./Std_Output/3-IDR/ \
        ./Scripts/IDR_STEP2.1_${day}${mark}.txt
  done ;
done


##--> STEP 2.2 : Peak calling on the two PR and the pooled replicates
for mark in ${marks[@]} ; do
 for day in ${days[@]} ; do
  for pr in REP1 REP2 ; do
        
        qsub \
            -l mem_rsvd=16G \
            -q ono \
            -cwd \
            -V \
            -N IDR_Step_2.2 \
            -e ./Std_Output/3-IDR/ \
            -o ./Std_Output/3-IDR/ \
            -b y macs2 callpeak callpeak --format BAMPE --gsize hs --keep-dup all -p $pvalPeaks --tempdir ./ -t ./Data/3-IDR/${day}${mark}_POOLED_${pr}_DEDUP_NOBLACKLIST_ONLYMAPPED.bam -n ./Data/3-IDR/${day}${mark}_POOLED_${pr}_MACS2
    
  done ;
 done ;
done

for mark in ${marks[@]} ; do
 for day in ${days[@]} ; do

       qsub \
            -l mem_rsvd=16G \
            -q ono \
            -cwd \
            -V \
            -N IDR_Step_2.2 \
            -e ./Std_Output/3-IDR/ \
            -o ./Std_Output/3-IDR/ \
            -b y macs2 callpeak callpeak --format BAMPE --gsize hs --keep-dup all -p $pvalPeaks --tempdir ./ -t ./Data/3-IDR/${day}${mark}_POOLED_DEDUP_NOBLACKLIST_ONLYMAPPED.bam -n ./Data/3-IDR/${day}${mark}_POOLED_MACS2
    
 done ;
done ;
```

### IDR 

```{bash, eval = FALSE}

##--> STEP 3.1 : Sort peaks by pval and select the first 100000
for mark in ${marks[@]} ; do
 for day in ${days[@]} ; do

   sort -k 8nr,8nr ./Data/3-IDR/${day}${mark}_POOLED_REP1_MACS2_peaks.narrowPeak | head -n 100000 | gzip -c > ./Data/3-IDR/${day}${mark}_POOLED_REP1_MACS2_peaks.regionPeak.gz
   sort -k 8nr,8nr ./Data/3-IDR/${day}${mark}_POOLED_REP2_MACS2_peaks.narrowPeak | head -n 100000 | gzip -c > ./Data/3-IDR/${day}${mark}_POOLED_REP2_MACS2_peaks.regionPeak.gz
   sort -k 8nr,8nr ./Data/3-IDR/${day}${mark}_Rep2_PR1_MACS2_peaks.narrowPeak | head -n 100000 | gzip -c > ./Data/3-IDR/${day}${mark}_Rep2_PR1_MACS2_peaks.regionPeak.gz
   sort -k 8nr,8nr ./Data/3-IDR/${day}${mark}_Rep2_PR2_MACS2_peaks.narrowPeak | head -n 100000 | gzip -c > ./Data/3-IDR/${day}${mark}_Rep2_PR2_MACS2_peaks.regionPeak.gz
   sort -k 8nr,8nr ./Data/3-IDR/${day}${mark}_Rep3_PR1_MACS2_peaks.narrowPeak | head -n 100000 | gzip -c > ./Data/3-IDR/${day}${mark}_Rep3_PR1_MACS2_peaks.regionPeak.gz
   sort -k 8nr,8nr ./Data/3-IDR/${day}${mark}_Rep3_PR2_MACS2_peaks.narrowPeak | head -n 100000 | gzip -c > ./Data/3-IDR/${day}${mark}_Rep3_PR2_MACS2_peaks.regionPeak.gz

 done ;
done;

for mark in ${marks[@]} ; do
 for day in ${days[@]} ; do
   for rep in ${reps[@]} ; do
    
    sort -k 8nr,8nr ./Data/3-PEAK-CALLING/MACS/FRAG_ALL/SHARP/${day}${mark}_${rep}_MACS2_FRAG_ALL_SHARP_peaks.narrowPeak | head -n 100000 | gzip -c > ./Data/3-IDR/${day}${mark}_${rep}_MACS2_peaks.regionPeak.gz
   
  done ;
 done ;
done;

for mark in ${marks[@]} ; do
 for day in ${days[@]} ; do
    
    sort -k 8nr,8nr ./Data/3-IDR/${day}${mark}_POOLED_MACS2_peaks.narrowPeak | head -n 100000 | gzip -c > ./Data/3-IDR/${day}${mark}_POOLED_MACS2_peaks.regionPeak.gz

 done ;
done;


##--> STEP 3.2 : Run the IDR

# Creation of batch script
for mark in ${marks[@]} ; do
 for day in ${days[@]} ; do
     
    echo "\
    #IDR ANALYSIS ON ORIGINAL REPLICATES 
    Rscript ./Data/3-IDR/batch-consistency-analysis.r ./Data/3-IDR/"${day}${mark}"_Rep2_MACS2_peaks.regionPeak.gz ./Data/3-IDR/"${day}${mark}"_Rep3_MACS2_peaks.regionPeak.gz -1 ./Data/3-IDR/"${day}${mark}"_Rep2_vs_Rep3 0 F p.value ;
    
    #IDR ANALYSIS ON SELF-PSEUDOREPLICATES
    Rscript ./Data/3-IDR/batch-consistency-analysis.r ./Data/3-IDR/"${day}${mark}"_Rep2_PR1_MACS2_peaks.regionPeak.gz ./Data/3-IDR/"${day}${mark}"_Rep2_PR2_MACS2_peaks.regionPeak.gz -1 ./Data/3-IDR/"${day}${mark}"_Rep2.1_vs_Rep2.2 0 F p.value ;
    Rscript ./Data/3-IDR/batch-consistency-analysis.r ./Data/3-IDR/"${day}${mark}"_Rep3_PR1_MACS2_peaks.regionPeak.gz ./Data/3-IDR/"${day}${mark}"_Rep3_PR2_MACS2_peaks.regionPeak.gz -1 ./Data/3-IDR/"${day}${mark}"_Rep3.1_vs_Rep3.2 0 F p.value ;
    
    #IDR ANALYSIS ON POOLED-PSEUDOREPLICATES
    Rscript ./Data/3-IDR/batch-consistency-analysis.r ./Data/3-IDR/"${day}${mark}"_POOLED_REP1_MACS2_peaks.regionPeak.gz ./Data/3-IDR/"${day}${mark}"_POOLED_REP2_MACS2_peaks.regionPeak.gz -1 ./Data/3-IDR/"${day}${mark}"_pooledPRep1_vs_pooledPRep2 0 F p.value" \
  >> ./Scripts/IDR_STEP3.2_${day}${mark}.txt
  
  done;
done;


# Run the analysis
for mark in ${marks[@]} ; do
 for day in ${days[@]} ; do
    qarray \
        -l mem_rsvd=8G \
        -q ono \
        -cwd \
        -V \
        -N IDR_Step_3.2 \
        -e ./Std_Output/3-IDR/ \
        -o ./Std_Output/3-IDR/ \
        ./Scripts/IDR_STEP3.2_${day}${mark}.txt
  done ;
done

##--> STEP 3.3 : Summarize the results

for mark in ${marks[@]} ; do
 for day in ${days[@]} ; do
     
  sigRep=.02 ; sigPRep=.02 ; sigPooledPrep=.005
  numPeaks_Rep1Rep2=$(awk -v s=$sigRep '$11 <= s {print $0}' ./Data/3-IDR/${day}${mark}_Rep2_vs_Rep3-overlapped-peaks.txt | wc -l)
  numPeaks_PRep11PRep12=$(awk -v s=$sigPRep '$11 <= s {print $0}' ./Data/3-IDR/${day}${mark}_Rep2.1_vs_Rep2.2-overlapped-peaks.txt | wc -l)
  numPeaks_PRep21PRep22=$(awk -v s=$sigPRep '$11 <= s {print $0}' ./Data/3-IDR/${day}${mark}_Rep3.1_vs_Rep3.2-overlapped-peaks.txt | wc -l)
  numPeaks_pooledPRep1PRep2=$(awk -v s=$sigPooledPrep '$11 <= s {print $0}' ./Data/3-IDR/${day}${mark}_pooledPRep1_vs_pooledPRep2-overlapped-peaks.txt | wc -l)
    
  echo $mark $day  
  echo $numPeaks_Rep1Rep2 $numPeaks_pooledPRep1PRep2 : reps vs pooled
  echo $numPeaks_PRep11PRep12 $numPeaks_PRep21PRep22 : preps

  if [ $numPeaks_Rep1Rep2 -gt $numPeaks_pooledPRep1PRep2 ] ; then opTh=$numPeaks_Rep1Rep2 ; else opTh=$numPeaks_pooledPRep1PRep2 ; fi

  echo choosen opTh $opTh 

  zcat ./Data/3-IDR/${day}${mark}_POOLED_MACS2_peaks.regionPeak.gz  | sort -k8nr,8nr | head -n $opTh > ./Data/3-IDR/${day}${mark}_MACS2_OPTIMAL_IDR.bed
  
 done;
done;

##--> Create a global .bed file with regions opened in a least one time point
for i in $(ls ./Data/3-IDR/ | grep _OPTIMAL_IDR.bed | grep -v MERGE | sed 's/_MACS2_OPTIMAL_IDR.bed//g')
do
    bedtools sort -i ./Data/3-IDR/"$i"_MACS2_OPTIMAL_IDR.bed > ./Data/3-IDR/"$i"_MACS2_OPTIMAL_IDR.bed.tmp
    mv ./Data/3-IDR/"$i"_MACS2_OPTIMAL_IDR.bed.tmp ./Data/3-IDR/"$i"_MACS2_OPTIMAL_IDR.bed
done

cat ./Data/3-IDR/D*ATAC_MACS2_OPTIMAL_IDR.bed | bedtools sort -i - | bedtools merge  -i - > ./Data/3-IDR/ATAC_IDR_COMBINE_SORT.bed

```

# Annotation

```{r}
##--> Load the GenomicRanges and rtracklayer package
library(GenomicRanges)
library(rtracklayer)

##--> Import annotation
annot_atac <- read.table("./results/MICROARRAY_ANNOTATION_BIOCONDUCTOR.txt", sep = "\t",
                         header = TRUE, stringsAsFactors = FALSE)

##--> Format
annot_atac <- annot_atac[-which(is.na(annot_atac$Chromosome) |
                                is.na(annot_atac$Start) |
                                is.na(annot_atac$Stop)),]

annot_atac <- GRanges(seqnames = paste("chr", annot_atac$Chromosome, sep = ""),
                      IRanges(start = annot_atac$Start, end = annot_atac$Stop),
                      mcols = data.frame(probes = as.character(annot_atac$Probe_ID),
                                         gene = as.character(annot_atac$HGNC_Symbol)))

##--> Liftover from hg38 to hg19
chain <- import.chain("./tools/hg38ToHg19.over.chain")
annot_chip <- unlist(liftOver(annot_atac, chain))
```

# Quality check

## Genome-wide correlation heatmap

### Preparing data

```{r}
##--> Load the GenomicRanges package
library(GenomicRanges)
library(GenomicAlignments)
library(BiocParallel)
library(rtracklayer)

##--> List alignment files
bam_atac <- list.files(path = "./data/ATAC-seq/",
                       pattern = ".bam$",
                       full = TRUE)

##---> Import the merge sets of reproducible ATAC peaks
peaks <- import("./data/ATAC-seq/GOLDSTANDARD_ATACSEQ_ALL_DAY_IDR.bed")
```

### Counting reads inside each peaks

```{r}
##--> Compute genome-wide coverage 
# Create a function to compute the coverage
summarizeOverlaps_batch <- function(bam) {
    count <- summarizeOverlaps(peaks, bam,ignore.strand = TRUE, inter.feature = FALSE)
    return(count)
}

# Run the counting in multi-thread
count_all <- bplapply(bam_atac,
                      summarizeOverlaps_batch, BPPARAM = MulticoreParam(bpprogressbar=TRUE))

# Transform the SummarizedExperiment object in numeric matrix
count_all_matrix <- do.call(cbind, lapply(count_all, assay))

# Change names of columns
colnames(count_all_matrix) <- gsub("RASOIS_ATACSEQ_|\\.bam", "", colnames(count_all_matrix)) 
```

### Normalization and debatching

```{r}
#--> Load libraries
library(RUVSeq)

#--> Quantile normalization
qn_counts <- betweenLaneNormalization(count_all_matrix, which = "full")

#--> Debatching
set <- newSeqExpressionSet(round(qn_counts))
differences <- matrix(data = c(1:18), ncol = 3, byrow = TRUE)
set_RUVs <- RUVs(x = set, cIdx = row.names(set), k = 3, scIdx = differences)
qn_and_ruv_count <- normCounts(set_RUVs)
```

### Creating the pairwise correlation matrix

```{r}
##--> Calculate pariwise Pearson's correlation before / after debatching
cor_count_all_matrix_qn  <- cor(qn_counts, method = "pearson")
cor_count_all_matrix_ruv <- cor(qn_and_ruv_count, method = "pearson")
```

### Drawing the correlation heatmap

```{r, fig.cap= "Pair-wise Pearson's correlation heatmap for genome-wide histone modification ATAC-seq", fig.height=15}
##--> Load the packages
library(pheatmap)
library(colorRamps)
library(gridExtra)

##--> Draw heatmap
P1 <- pheatmap(cor_count_all_matrix_qn,
         clustering_distance_rows = "correlation",
         clustering_distance_cols = "correlation",
         clustering_method = "ward.D2", silent = TRUE,
         main = "Pair-wise Pearson's correlation heatmap - Quantile normalization",
         color = matlab.like(c(1000)))

P2 <- pheatmap(cor_count_all_matrix_ruv,
         clustering_distance_rows = "correlation",
         clustering_distance_cols = "correlation",
         clustering_method = "ward.D2", silent = TRUE,
         main = "Pair-wise Pearson's correlation heatmap - Quantile normalization + Debatching",
         color = matlab.like(c(1000)))

do.call(grid.arrange,list(P1[[4]],P2[[4]]))
```

## Saturation analysis

As a prerequisite for saturation analysis, we do the following:

 * cut the genome in bins of fixed size (i.e. 1kb),
 * count the number of reads aligned on each of these bins for each alignment files,
 * remove bins on which no alignment is reported in all of the alignment files (otherwise, it will artificially increase the correlation),

We will then use the raw count matrix `count_all_matrix` that includes bins of the genome where no read aligns. Based on these counts, for each library independently, we then interpolate a species accumulation curve using the package `r CRANpkg("preseqR")`.This allows to predict the molecular complexity of sequencing libraries. The idea is to use the information gained from the number of times each 1kb is observed to predict the number of new, currently unobserved, bins that will be gained from additional sequencing.

### Preparing data

```{r}
##--> Load the GenomicRanges package
library(preseqR)

##--> List alignment files
bam_atac <- list.files(path = "./data/ATAC-seq/",
                       pattern = ".bam$",
                       full = TRUE)

library(BSgenome.Hsapiens.UCSC.hg19)
genome <- BSgenome.Hsapiens.UCSC.hg19

##--> Use the tileGenome function to tile the genome
tile <- tileGenome(seqlengths(genome), tilewidth = 1000)
```

### Counting reads inside each bin

```{r}
##--> Compute genome-wide coverage 
# Create a function to compute the coverage
summarizeOverlaps_batch <- function(bam) {
    count <- summarizeOverlaps(tile, bam, ignore.strand = TRUE, inter.feature = FALSE)
    return(count)
}

# Run the counting in multi-thread
count_all <- bplapply(bam_atac, summarizeOverlaps_batch, BPPARAM = MulticoreParam())

# Transform the SummarizedExperiment object in numeric matrix
count_all_matrix <- do.call(cbind, lapply(count_all, assay))

# Change names of columns
colnames(count_all_matrix) <- gsub("RASOIS_ATACSEQ_|\\.bam", "", colnames(count_all_matrix)) 

##--> Indentifying and removing bins with no read
# Sum across lines of the count matrix
zero <- apply(count_all_matrix, 1, sum)

# Remove lines for which the sum == 0
count_all_matrix_clean <- count_all_matrix[-which(zero==0),]
```

```{r, fig.cap= "Species accumulation curves for genome-wide histone modification ATAC-seq", fig.height=10}
##--> Load the preseqR package
library(preseqR)

##--> Select data for the saturation analysis
saturation <- count_all_matrix_clean

##--> Create the object used to output the results for each saturation analysis
saturation_res <- data.frame()

##--> Runing saturation analysis for each library
for (i in 1:ncol(saturation)) {
  table <- table(saturation[,i])
  test <- preseqR.interpolate.distinct(1000000, as.matrix(data.frame(cbind(as.double(names(table)),table)))[-1,])
  test <- data.frame(test)
  test$Name <- rep(colnames(saturation)[i],nrow(test))
  saturation_res <- rbind(saturation_res, test)
}

##--> Reordering the output for clarity
saturation_res$Day <- factor(gsub(".*_ATACSEQ_|_REP[123]", "", saturation_res$Name),  levels = c("T0", "24H", "48H",
                                                                                                 "72H",  "96H", "144H"))
saturation_res$Rep <- as.factor(gsub(".*_", "", saturation_res$Name))

##--> Ploting results
library(ggplot2)
saturation_plot <- ggplot(saturation_res, aes(x = sample.size, y = interpolation)) +
  geom_point(aes(fill = Rep), alpha = 0.7, colour="black",pch=21, size=1) +
  facet_wrap(~ Day) +
  scale_x_continuous(name = "Number of mapped reads") +
  scale_y_continuous(name = "Number of peaks covered") +
  ggtitle("Mapped-bins accumulation curves for mapped reads in each libraries") +
  theme_bw()
saturation_plot
```

## Genome-wide profiles around TSS

Before starting this analysis, we need to retrieve the coordinates of TSS on a data-base such as UCSC or Biomart.

### Preparing the data 

```{r}
##--> Load the biomeRt and the biovizBase packages
library(biomaRt)
library(biovizBase)

##--> Connexion to BioMart hg19 (aka GRCh37)
ensembl <- useMart(biomart="ENSEMBL_MART_ENSEMBL",
                   host="grch37.ensembl.org",
                   path="/biomart/martservice",
                   dataset="hsapiens_gene_ensembl")

##--> Retrive the coordinates of TSSs
tss <- getBM(mart = ensembl,
             attributes = c("chromosome_name",
                            "transcription_start_site",
                            "hgnc_symbol",
                            "strand"))

##--> Format the chromosome name column (to match the chromosome names in alignement files)
set.seed(1234)
tss <- tss[grep("^[1-9]{1,2}$", tss$chromosome_name),]
tss <- tss[sample(1:nrow(tss), 10000),]
tss$chromosome_name <- paste("chr", tss$chromosome_name, sep = "")

##--> Creating a GRange containing the information related to TSSs
# And add a +/- 1kb window
tss_gr <- GRanges(seqnames = tss$chromosome_name,
                  ranges = IRanges(start = tss$transcription_start_site - 500,
                                   end = tss$transcription_start_site + 500,
                                   names = tss$hgnc_symbol),
                  strand = tss$strand)

##--> For each TSS +/- 500bp, tile the window in 100 bins
tss_split <- tile(tss_gr, n = 100)
tss_split <- flatGrl(tss_split) 
```

Now that the object containing information related to TSSs is formated, we will first list the alignement files we are interested in, and count the reads overlapping each tiles consituting each TSS window.

### Counting reads inside each bin

```{r}
##--> Create a function to compute the coverage
summarizeOverlaps_tss_batch <- function(bam) {
    count <- summarizeOverlaps(tss_split,
                               bam,
                               inter.feature = FALSE,
                               fragments = FALSE,
                               singleEnd = FALSE,
                               param = ScanBamParam(which = tss_gr))
    return(count)
}

##--> Count reads in bins
count_tss <- bplapply(bam_atac, summarizeOverlaps_tss_batch, BPPARAM = MulticoreParam())

##--> Transform the SummarizedExperiment object in numeric matrix
count_tss_matrix <- do.call(cbind, lapply(count_tss, assay))

##--> Change names of columns
colnames(count_tss_matrix) <- gsub("RAS_OIS_ATACSEQ_|\\.bam", "", colnames(count_tss_matrix)) 
```

### Formating output and plot

```{r, fig.cap= "TSS metaprofiles for histone modification ChIP-seq", fig.height=7}
day  <- factor(gsub(".*_ATACSEQ_|_REP[123]", "", colnames(count_tss_matrix)), 
               levels = c("T0", "24H", "48H","72H",  "96H", "144H"))

##--> Create a list of count tables (one per sample)
count_tss_list <- lapply(split(count_tss_matrix,
                               rep(1:ncol(count_tss_matrix),
                                   each = nrow(count_tss_matrix))),
                         function(x) {matrix(x, ncol=100, byrow = TRUE)})
names(count_tss_list) <- colnames(count_tss_matrix)

##--> Take the gene strand into account
count_tss_list <- lapply(count_tss_list, function(x) {
  x[which(strand(tss_gr) == "+"),c(1:100)] <- x[which(strand(tss_gr) == "+"), c(1:100)]
  x[which(strand(tss_gr) == "-"),c(1:100)] <- x[which(strand(tss_gr) == "-"), c(100:1)]
})

##--> Compute a CPM-normalized average signal
count_tss_list <- lapply(count_tss_list, function(x) {
  #x <- x / sum(x) * 1000000
  x <- colMeans(x)
})


##--> Format
count_tss_plot <- reshape2::melt(count_tss_list)
count_tss_plot$Dist_to_TSS <- seq(from = -500, to = 500, length.out = 101)[-1]
count_tss_plot$Time  <- factor(gsub(".*_ATACSEQ_|_REP[123]", "", count_tss_plot$L1), 
                                levels = c("T0", "24H", "48H","72H",  "96H", "144H"))
count_tss_plot$Replicate <- factor(gsub("RASOIS_.*_", "", count_tss_plot$L1), 
                                levels = c("REP1", "REP2", "REP3"))
colnames(count_tss_plot) <- c("RPM", "Bam", "Dist_to_TSS", "Time", "Replicate")

##--> Create Plots
ggplot(count_tss_plot, aes(x = Dist_to_TSS, y = RPM, group = Replicate)) +
  geom_line(aes(color = Replicate)) +
  facet_wrap(~Time) +
  scale_x_continuous(name = "Distance to TSS (bp)") +
  scale_y_continuous(name = "Average read per million mapped reads (N)") +
  ggtitle("TSS metaprofiles for histone modification ChIP-seq") +
  theme_bw()

```

## Fragment length distribution

```{r, fig.cap= "Fragment length distribution profiles"}
##--> Load libraries
library(BiocParallel)
library(Rsamtools)
library(ggplot2)

##--> Create a function to get fragment length from .bam files
get_fl <- function(bam) {
        fl <- scanBam(file = bam, 
                     index = bam,
                     use.names = TRUE,
                     param = ScanBamParam(what = c("isize")), which = peaks)
    return(fl)
}

##--> Count reads in bins
fl <- bplapply(bam_atac, get_fl, BPPARAM = MulticoreParam())

##--> Generate counting tables
fl_table <- lapply(fl, function(x) {table(abs(unlist(x)))})

##--> Reshape and scale
names(fl_table) <- gsub("RAS_OIS_ATACSEQ_|\\.bam", "", colnames(count_tss_matrix)) 
fl_table_2 <- lapply(fl_table, function(x) {(x / sum(x)) * 10^6})
fl_table_2 <- reshape2::melt(fl_table_2)

##--> Format
fl_table_2$Time  <- factor(gsub(".*_ATACSEQ_|_REP[123]", "", fl_table_2$L1), 
                                levels = c("T0", "24H", "48H","72H",  "96H", "144H"))
fl_table_2$Replicate <- factor(gsub("RASOIS_.*_", "", fl_table_2$L1), 
                                levels = c("REP1", "REP2", "REP3"))
colnames(fl_table_2) <- c("FL", "Frequence", "Bam", "Time", "Replicate")

##--> Create Plots
ggplot(fl_table_2, aes(x= FL, y = (Frequence), group = Time)) +
  geom_line(aes(color = Time)) +
  scale_color_manual(values = rainbow(n=12)[c(1,7:12)]) +
  scale_x_continuous(limits = c(1,500)) + 
  xlab("Fragment length") +
  ylab("Counts") +
  facet_wrap(~Replicate)

```

# Nucleosome positioning 

Preprocessed ATAC-seq data were first down-sampled to reach for each library the minimal number of reads obtained in each batch across time-points and replicates. Down-sampled replicates corresponding to a given time-point were then merged. These down-sampled and merged data were then used in combination with chromatin accessible regions defined before as the input to [NucleoATAC v0.3.1](https://github.com/GreenleafLab/NucleoATAC/releases) to define nucleosome density inside accessible regions.

## Merging replicates and downsampling

```{bash, eval = FALSE}

##--> Merge bam files from replicates

days=(D0 D1 D2 D3 D4 D6)
for day in ${days[@]}
do
    qsub \
     -q ono \
     -cwd \
     -V \
     -pe thread 8 \
     -b y samtools merge -@ 8 -h ./Data/1-NOBLACKLIST_Bam/Header.txt \
            ./Data/1-NOBLACKLIST_Bam/${day}ATAC_MERGE_123_DEDUP_NOBLACKLIST.bam \
            ./Data/1-NOBLACKLIST_Bam/${day}ATAC_Rep1_DEDUP_NOBLACKLIST_SORT.bam \
            ./Data/1-NOBLACKLIST_Bam/${day}ATAC_Rep2_DEDUP_NOBLACKLIST_SORT.bam \
            ./Data/1-NOBLACKLIST_Bam/${day}ATAC_Rep3_DEDUP_NOBLACKLIST_SORT.bam
done

##--> Subsample merged files

nlinesD0=$(samtools idxstats ./Data/1-NOBLACKLIST_Bam/D0ATAC_MERGE_123_DEDUP_NOBLACKLIST.bam | awk '{s+=$3} END {print s}');
nlinesD1=$(samtools idxstats ./Data/1-NOBLACKLIST_Bam/D1ATAC_MERGE_123_DEDUP_NOBLACKLIST.bam | awk '{s+=$3} END {print s}');
nlinesD2=$(samtools idxstats ./Data/1-NOBLACKLIST_Bam/D2ATAC_MERGE_123_DEDUP_NOBLACKLIST.bam | awk '{s+=$3} END {print s}');
nlinesD3=$(samtools idxstats ./Data/1-NOBLACKLIST_Bam/D3ATAC_MERGE_123_DEDUP_NOBLACKLIST.bam | awk '{s+=$3} END {print s}');
nlinesD4=$(samtools idxstats ./Data/1-NOBLACKLIST_Bam/D4ATAC_MERGE_123_DEDUP_NOBLACKLIST.bam | awk '{s+=$3} END {print s}');
nlinesD6=$(samtools idxstats ./Data/1-NOBLACKLIST_Bam/D6ATAC_MERGE_123_DEDUP_NOBLACKLIST.bam | awk '{s+=$3} END {print s}');
nlinesMin=$(echo "$nlinesD0"; echo "$nlinesD1"; echo "$nlinesD2"; echo "$nlinesD3"; echo  "$nlinesD4"; echo  "$nlinesD6");
nlinesMin=$(echo $nlinesMin |sed 's/\s/\n/g' | sort -n| head -n1)

probD0=$(echo $nlinesMin $nlinesD0 | awk '{print $1/$2}')
probD1=$(echo $nlinesMin $nlinesD1 | awk '{print $1/$2}')
probD2=$(echo $nlinesMin $nlinesD2 | awk '{print $1/$2}')
probD3=$(echo $nlinesMin $nlinesD3 | awk '{print $1/$2}')
probD4=$(echo $nlinesMin $nlinesD4 | awk '{print $1/$2}')
probD6=$(echo $nlinesMin $nlinesD6 | awk '{print $1/$2}')
prob=($probD0 $probD1 $probD2 $probD3 $probD4 $probD6)


# Create a batch script
days=(D0 D1 D2 D3 D4 D6)
i=0
for day in ${days[@]}
do
    echo "java -jar -Xmx16g ~/bin/Tools/PicardTools_v1.130/picard.jar \
              DownsampleSam \
              PROBABILITY=${prob[i]} \
              I=./Data/1-NOBLACKLIST_Bam/"$day"ATAC_MERGE_123_DEDUP_NOBLACKLIST.bam \
              O=./Data/1-NOBLACKLIST_Bam/"$day"ATAC_MERGE_123_DEDUP_NOBLACKLIST_DOWNSAMPLED.bam" \
        >> ./Scripts/DOWNSAMPLING.txt
    i=$i+1
done


# Run the analysis        
qarray \
    -q ono \
    -cwd \
    -V \
    -N Downsampling \
    -e ./Std_Output/1-Downsampling/ \
    -o ./Std_Output/1-Downsampling/ \
    ./Scripts/DOWNSAMPLING.txt   
```


## Nucleosome calling

```{bash, eval = FALSE}
##--> Combine and / or Extend intervals in MAC2 IDR peaks

bedtools slop -b 200 -i ./Data/3-IDR/ATAC_IDR_COMBINE_SORT.bed -g ~/save/Outils/hg19_chrom.sizes | \
awk -f ~/bin/Scripts/combine_extended_bed.awk \
> ./Data/3-PEAK-CALLING/NucleoATAC/ATAC_IDR_COMBINE_SORT_EXTENDED_200BP.bed

cat ./Data/3-IDR/ATAC_IDR_COMBINE_SORT.bed | \
awk -f ~/bin/Scripts/combine_extended_bed.awk \
> ./Data/3-PEAK-CALLING/NucleoATAC/ATAC_IDR_COMBINE_SORT.bed

##--> NucleoATAC workflow

# Creation of batch script
for i in $(ls ./Data/1-NOBLACKLIST_Bam/ |grep _DOWNSAMPLED.bam$ | grep -v FRAG | grep -v Rep1 | sed 's/_MERGE_123_DEDUP_NOBLACKLIST_DOWNSAMPLED.bam//g')
do
    echo "\
    nucleoatac occ \
      --cores 8 \
      --bed ./Data/3-PEAK-CALLING/NucleoATAC/ATAC_IDR_COMBINE_SORT_EXTENDED_200BP.bed \
      --bam ./Data/1-NOBLACKLIST_Bam/"$i"_MERGE_123_DEDUP_NOBLACKLIST_DOWNSAMPLED.bam \
      --fasta /local/databases/index/samtools/hg19/0.1.19/hg19.fa \
      --out ./Data/3-PEAK-CALLING/NucleoATAC/"$i"_MERGE_123;
    
    nucleoatac vprocess \
      --size ./Data/3-PEAK-CALLING/NucleoATAC/"$i"_MERGE_123.nuc_dist.txt \
      --out ./Data/3-PEAK-CALLING/NucleoATAC/"$i"_MERGE_123;
    
    nucleoatac nuc \
      --cores 8 \
      --sizes ./Data/3-PEAK-CALLING/NucleoATAC/"$i"_MERGE_123.nuc_dist.txt \
      --occ_track ./Data/3-PEAK-CALLING/NucleoATAC/"$i"_MERGE_123.occ.bedgraph.gz \
      --vmat ./Data/3-PEAK-CALLING/NucleoATAC/"$i"_MERGE_123.VMat \
      --bed ./Data/3-PEAK-CALLING/NucleoATAC/ATAC_IDR_COMBINE_SORT_EXTENDED_200BP.bed \
      --bam ./Data/1-NOBLACKLIST_Bam/"$i"_MERGE_123_DEDUP_NOBLACKLIST_DOWNSAMPLED.bam \
      --fasta /local/databases/index/samtools/hg19/0.1.19/hg19.fa \
      --out ./Data/3-PEAK-CALLING/NucleoATAC/"$i"_MERGE_123;
      
    nucleoatac merge \
      --occpeaks ./Data/3-PEAK-CALLING/NucleoATAC/"$i"_MERGE_123.occpeaks.bed.gz \
      --nucpos ./Data/3-PEAK-CALLING/NucleoATAC/"$i"_MERGE_123.nucpos.bed.gz;
      
    nucleoatac nfr \
      --cores 8 \
      --bed ./Data/3-PEAK-CALLING/NucleoATAC/ATAC_IDR_COMBINE_SORT_EXTENDED_200BP.bed \
      --occ_track ./Data/3-PEAK-CALLING/NucleoATAC/"$i"_MERGE_123.occ.bedgraph.gz \
      --calls ./Data/3-PEAK-CALLING/NucleoATAC/"$i"_MERGE_123.occpeaks.bed.gz \
      --bam ./Data/1-NOBLACKLIST_Bam/"$i"_MERGE_123_DEDUP_NOBLACKLIST_DOWNSAMPLED.bam \
      --fasta /local/databases/index/samtools/hg19/0.1.19/hg19.fa --out ./Data/3-PEAK-CALLING/NucleoATAC/"$i"_MERGE_123"\
    >> ./Scripts/NucleoATAC_$i.txt
done

for i in $(ls ./Data/1-NOBLACKLIST_Bam/ |grep _DOWNSAMPLED.bam$ | grep -v FRAG | grep -v Rep1 | sed 's/_MERGE_123_DEDUP_NOBLACKLIST_DOWNSAMPLED.bam//g')
do
    qsub \
        -q ono \
        -cwd \
        -V \
        -N Nucleo_"$i" \
        -pe thread 8 \
        -e ./Std_Output/3-NucleoATAC/ \
        -o ./Std_Output/3-NucleoATAC/ \
        ./Scripts/NucleoATAC_"$i".txt
done

```

# Transcription factor footprinting

[PIQ](https://bitbucket.org/thashim/piq-single/src) was used to predict transcription factor binding sites from the genome sequence on down-sampled ATAC-seq alignments. For each motif, we retained only binding sites that were within the reproducible ATAC-seq peaks and passed the default purity cut-off (70%).

## Footprinting wit PIQ

```{bash, eval = FALSE}

days=(D0 D1 D2 D3 D4 D6)

##--> STEP 1 : Convert BAM to internal binary format (does not depend on choice of motif).
# [arg1]: file with basic parameters,
# [arg2]: name of the output, 
# [arg3]: name of the input

# For each time points
for day in ${days[@]}
do
    qsub -q ono -cwd -l mem_rsvd=128G -V -b y Rscript ~/bin/Tools/PIQ/pairedbam2rdata.r \
                                         ~/bin/Tools/PIQ/common.r \
                                         ./Data/5-FOOTPRINTING/PIQ/"$day"ATAC_MERGE_123_DOWNSAMPLED_PIQ.RData \
                                         ./Data/1-NOBLACKLIST_Bam/"$day"ATAC_MERGE_123_DEDUP_NOBLACKLIST_DOWNSAMPLED.bam
done


##--> STEP 2 : Generate the PWM hits across genome (does not depend on choice of BAM).
# [arg1]: file with basic parameters,
# [arg2]: file containing the pwms, 
# [arg3]: indice of the motif of interest on the file containing pwms

# Create a batch file
for i in $(seq 519)
do
    echo "Rscript ~/bin/Tools/PIQ/pwmmatch.exact.r \
          ~/bin/Tools/PIQ/common.r \
          ~/bin/Tools/PIQ/pwms/jaspar_eukaryotes_519_TF.txt "$i" \
          ./Data/5-FOOTPRINTING/PIQ/Motif_Matches/" \
    >> ./Scripts/PIQ_PWM_MATCH.txt
done

# Run the analysis
qarray \
        -q ono \
        -cwd \
        -V \
        -N PIQ_PWM \
        -e ./Std_Output/5-PIQ/ \
        -o ./Std_Output/5-PIQ/ \
        ./Scripts/PIQ_PWM_MATCH.txt


##--> STEP 3 : Combine BAM + PWM  (depends on choice of BAM and PWM).
# [arg1]: file with basic parameters,
# [arg2]: repository containing the motif matches of the motif of interest,
# [arg3]: temporty repository, 
# [arg4]: output repository, 
# [arg5]: converted bam file

for day in ${days[@]}
do
    mkdir -p ./Data/5-FOOTPRINTING/PIQ/Tmp/"${day}"
    mkdir -p ./Data/5-FOOTPRINTING/PIQ/ATAC_Matches/"${day}"
done
 
# Create a batch file
for i in $(seq 519); do
    echo "Rscript ~/bin/Tools/PIQ/permark.r \
        ~/bin/Tools/PIQ/common.r \
        ./Data/5-FOOTPRINTING/PIQ/Motif_Matches/ \
        ./Data/5-FOOTPRINTING/PIQ/Tmp/ALL_ATAC/ \
        ./Data/5-FOOTPRINTING/PIQ/ATAC_Matches/ALL_ATAC/ \
        ./Data/5-FOOTPRINTING/PIQ/ALL_ATAC_PIQ_FIFTY_PERCENT.RData "$i"" \
        >> ./Scripts/PIQ_TRAIN.txt
done

# Run the analysis
qarray \
        -q ono \
        -cwd \
        -V \
        -N PIQ_TRAIN \
        -e ./Std_Output/5-PIQ/ \
        -o ./Std_Output/5-PIQ/ \
        ./Scripts/PIQ_TRAIN.txt

##--> STEP 4 : Combine BAM + PWM to TRAIN (depends on choice of BAM and PWM).
# [arg1]: file with basic parameters, 
# [arg2]: repository containing the motif matches of the motif of interest,
# [arg3]: temporty repository, 
# [arg4]: output repository, 
# [arg5]: converted bam file use for training
# [arg5]: converted bam file use for testing

# Create a batch file

for day in ${days[@]}; do
    for i in $(seq 519); do
        echo "Rscript ~/bin/Tools/PIQ/permark.r \
            ~/bin/Tools/PIQ/common.r \
            ./Data/5-FOOTPRINTING/PIQ/Motif_Matches/ \
            ./Data/5-FOOTPRINTING/PIQ/Tmp/${day}/ \
            ./Data/5-FOOTPRINTING/PIQ/ATAC_Matches/${day}/ \
            ./Data/5-FOOTPRINTING/PIQ/"$day"ATAC_MERGE_123_DOWNSAMPLED_PIQ.RData "$i"" \
        >> ./Scripts/PIQ_TEST.txt
    done
done

# Run the analysis
qarray \
        -q ono \
        -cwd \
        -V \
        -N PIQ_TEST \
        -e ./Std_Output/5-PIQ/ \
        -o ./Std_Output/5-PIQ/ \
        ./Scripts/PIQ_TEST.txt

##--> Rename output to remove space in the name
for day in ${days[@]}; do
    echo $day
    find ./Data/5-FOOTPRINTING/PIQ/ATAC_Matches/"$day"/ -type f -name "* *" -exec bash -c 'mv "$0" "${0// /_}"' {} \;
done

##--> Convert .csv file with true calls in .bed
for day in ${days[@]}; do
    echo $day
    for i in $(ls --color=never ./Data/5-FOOTPRINTING/PIQ/ATAC_Matches/"$day"/ | grep "chropen.txt"); do
            
        mark1=$(echo $i | sed 's/-chropen.txt//g')
        cat ./Data/5-FOOTPRINTING/PIQ/ATAC_Matches/"$day"/"$mark1"-calls.csv | sed 's/"//g'  | sed 's/,/\t/g' |\
        grep -vP '^\t' | awk '{OFS = "\t"};{print $2, $3, $3, $4, $5, $6, $7}' \
        > ./Data/5-FOOTPRINTING/PIQ/ATAC_Matches/Bilan_Call/PIQ_CALL_"$day"_"$mark1".bed
        
    done
done

##--> Intersect with the global IDR peak sets
for day in ${days[@]}; do

    echo $day
    
    for i in $(ls --color=never ./Data/5-FOOTPRINTING/PIQ/ATAC_Matches/"$day"/ | grep "chropen.txt"); do
    
        i=$(echo $i | sed 's/\s/_/g')
        mark1=$(echo $i | sed 's/-chropen.txt//g')
        intersectBed \
            -a ./Data/5-FOOTPRINTING/PIQ/ATAC_Matches/Bilan_Call/PIQ_CALL_"$day"_"$mark1".bed \
            -b ./Data/3-PEAK-CALLING/NucleoATAC/ATAC_IDR_COMBINE_SORT.bed \
            -wa -u \
            > ./Data/5-FOOTPRINTING/PIQ/ATAC_Matches/Bilan_Call/PIQ_CALL_IDR_"$day"_"$mark1".bed
            
    done
done

##--> Summarize the results in a global matrix

for i in $(ls ./Data/5-FOOTPRINTING/PIQ/ATAC_Matches/D0/ | grep "chropen.txt"); do

    mark1=$(echo $i | sed 's/-chropen.txt//g')
    mark2=$(echo $mark1 | sed 's/.*-//g')
    
    open_D0=$(cat ./Data/5-FOOTPRINTING/PIQ/ATAC_Matches/D0/"$i" |sed 's/,/\t/g' |cut -f 2)
    nsig_D0=$(cat ./Data/5-FOOTPRINTING/PIQ/ATAC_Matches/Bilan_Call/PIQ_CALL_IDR_D0_"$mark1".bed |wc -l)
    
    open_D1=$(cat ./Data/5-FOOTPRINTING/PIQ/ATAC_Matches/D1/"$i" |sed 's/,/\t/g' |cut -f 2)
    nsig_D1=$(cat ./Data/5-FOOTPRINTING/PIQ/ATAC_Matches/Bilan_Call/PIQ_CALL_IDR_D1_"$mark1".bed |wc -l)
    
    open_D2=$(cat ./Data/5-FOOTPRINTING/PIQ/ATAC_Matches/D2/"$i" |sed 's/,/\t/g' |cut -f 2)
    nsig_D2=$(cat ./Data/5-FOOTPRINTING/PIQ/ATAC_Matches/Bilan_Call/PIQ_CALL_IDR_D2_"$mark1".bed |wc -l)
    
    open_D3=$(cat ./Data/5-FOOTPRINTING/PIQ/ATAC_Matches/D3/"$i" |sed 's/,/\t/g' |cut -f 2)
    nsig_D3=$(cat ./Data/5-FOOTPRINTING/PIQ/ATAC_Matches/Bilan_Call/PIQ_CALL_IDR_D3_"$mark1".bed |wc -l)
    
    open_D4=$(cat ./Data/5-FOOTPRINTING/PIQ/ATAC_Matches/D4/"$i" |sed 's/,/\t/g' |cut -f 2)
    nsig_D4=$(cat ./Data/5-FOOTPRINTING/PIQ/ATAC_Matches/Bilan_Call/PIQ_CALL_IDR_D4_"$mark1".bed |wc -l)
    
    open_D6=$(cat ./Data/5-FOOTPRINTING/PIQ/ATAC_Matches/D6/"$i" |sed 's/,/\t/g' |cut -f 2)
    nsig_D6=$(cat ./Data/5-FOOTPRINTING/PIQ/ATAC_Matches/Bilan_Call/PIQ_CALL_IDR_D6_"$mark1".bed |wc -l)
    
    echo "$mark1 $open_D0 $nsig_D0 $open_D1 $nsig_D1 $open_D2 $nsig_D2 $open_D3 $nsig_D3 $open_D4 $nsig_D4 $open_D6 $nsig_D6" >> ./Data/5-FOOTPRINTING/PIQ/ATAC_Matches/BILAN_ALL_ATAC_2.txt
    
done
    
##--> Generate a global .bed file for each time point

for day in ${days[@]}; do

    echo $day;
    
    for i in $(ls --color=never ./Data/5-FOOTPRINTING/PIQ/ATAC_Matches/"$day"/ | grep "chropen.txt"); do
    
      mark1=$(echo $i | sed 's/-chropen.txt//g') ;
      mark2=$(echo $mark1 | sed 's/.*-//g') ;
    
      cat ./Data/5-FOOTPRINTING/PIQ/ATAC_Matches/Bilan_Call/PIQ_CALL_"$day"_"$mark1".bed \
      | awk -v mark=$mark2 '{OFS = "\t"};{print $1, $2, $3, mark, $4, $5, $6, $7}' \
       >> ./Data/5-FOOTPRINTING/PIQ/ATAC_Matches/Bilan_Call/PIQ_CALL_ALL_TF_"$day".bed.tmp ;
    
    done
    
    bedtools sort -i ./Data/5-FOOTPRINTING/PIQ/ATAC_Matches/Bilan_Call/PIQ_CALL_ALL_TF_"$day".bed.tmp > \
      ./Data/5-FOOTPRINTING/PIQ/ATAC_Matches/Bilan_Call/PIQ_CALL_ALL_TF_"$day".bed
    rm ./Data/5-FOOTPRINTING/PIQ/ATAC_Matches/Bilan_Call/PIQ_CALL_ALL_TF_"$day".bed.tmp ;
    bgzip ./Data/5-FOOTPRINTING/PIQ/ATAC_Matches/Bilan_Call/PIQ_CALL_ALL_TF_"$day".bed
    tabix ./Data/5-FOOTPRINTING/PIQ/ATAC_Matches/Bilan_Call/PIQ_CALL_ALL_TF_"$day".bed.gz ;
    
done


##--> Compute statistics

days=(D0 D1 D2 D3 D4 D6)

for day in ${days[@]}; do
    for i in $(cat ~/bin/Tools/PIQ/pwms/jaspar_eukaryotes_519_TF.txt | grep ">" | sed 's/_.*//g' |sed 's/>//g'); do
        echo "Rscript ~/bin/Scripts/PIQ_COI.R $i $day" >> ./Scripts/PIQ_COI.txt ;
    done
done

qarray \
    -q ono \
    -cwd \
    -V \
    -l mem_rsvd=32G \
    -N PIQ_COI \
    -e ./Std_Output/5-PIQ/ \
    -o ./Std_Output/5-PIQ/ \
    ./Scripts/PIQ_COI.txt
    
    
```

## Experimental validation

To compare PIQ prediction with RELA, JUN and FOSL2 ChIP-seq data, we first used the approach suggested in (Sherwood et al., 2014), computing how many of the total ChIP-seq peaks are overlapping with any potential factor motif (since ChIP-Seq peaks can result from co-factor binding, and methods such as DGF are factor agnostic). We then used a more sophisticated approach aiming at correlating the ChIP-seq signal intensity with the bound / unbound status at PWM matches. For a given transcription factor (RELA, JUN or FOSL2), we first considered all the PWM matches located inside ATAC-seq reproducible peaks, we selected all the PWM matches assigned with a purity score > 0.7 (the threshold used to define bound instances), and then randomly selected 3 times more PWM matches assigned with a purity score < 0.7 (considered as unbound instances) to obtain a global set containing 25% / 75% of bound / unbound instances for each TF. The selected regions were extended up to 2kb (1kb in each direction, from the middle of the match), and the 2kb intervals were binned in 100 20bp windows. We computed the normalized ChIP-seq and ATAC-seq signal inside each bin. The windows were finally ranked according to the summed ChIP-seq signal in the 10 most central bins (200bp). We finally run a functional enrichment analysis to assess whether bound / unbound PWM matches were enriched / depleted along this ranking and computed the enrichment score (ES, positive when bound instances are enriched for highest ChIP-seq signals, negative when unbound instances are depleted for highest ChIP-seq signals) and q-values which revealed the strength of the correlation.

### Overview of the data

The table below summarize the information relative to the TF ChIP-seq data anlayzed in the scope of this study. The information given are :

* *Time point* : T0 - before induction of RAS , 72h after induction or 144h induction),
* *Type* : type of the library (input, H3K4me1, H3K4me3, H3K27ac or H3K27me3),
* *Replicate* : there are two replciates per codition,
* *SRA* : Sequence Read Archive accession number,
* *Raw reads* : total number of reads for each library before any processing,
* *Clean reads* : number of reads remaining after quality filtering and adapter triming,
* *Uniquely mapped* : number of unambiguously mapped reads,
* *Non duplicated* : number of reads after filter for PCR / optical duplicates,
* *Not in BLR* : number of reads mapped outside hg19 ENCODE blacklisted regions (BLR),
* *% kept* : percentage of reads kept for subsequent analysis after filtering,
* *% duplicates* : percentage of PCR / optical duplicates,
* *% BLR* : percentage of reads mapped in BLR.

```{r, results= "asis", echo = FALSE}
library(dplyr)
library(knitr)
library(kableExtra)
info_chipseq <- read.table("./data/ChIP-seq/ChIP-seq_TF_data_info.txt", header = TRUE, sep ="\t")

kable(info_chipseq, ,
              col.names = c("Time point", "Type", "Replicate", "SRA",
                             "Raw reads (N)", "Clean reads (N)", "Uniquely mapped (N)",
                              "Non duplicated (N)", "Not in BLR (N)", "% kept",
                              "% duplicates", "% in BLR")) %>%
kable_styling("striped", full_width = TRUE, font_size = 9, position = "left") %>%
  collapse_rows(columns = 1:2, valign = "top")

```


### Preprocessing of TF ChIP-seq data

All the TF ChIP-seq data were processed using the workflow written in *bash* describe in the Appendix 2 and ran on a Unix cluster driven by Sun Gride Engine. The different steps we described bellow are :

* quality check on raw reads with [fastQC 0.11.5](https://www.bioinformatics.babraham.ac.uk/projects/fastqc/),
* read trimming and cleaning with [fastq-mcf 1.04.803](https://github.com/ExpressionAnalysis/ea-utils/tree/wiki),
* check for contaminations with [fastq-screen 0.4.4](https://www.bioinformatics.babraham.ac.uk/projects/fastq_screen/),
* alignment with [bowtie 1.1.1](http://bowtie-bio.sourceforge.net/index.shtml),
* de-duplication with [PicardTools 1.130](https://broadinstitute.github.io/picard/),
* filtering of reads aligned in [Encode blaclisted regions](https://sites.google.com/site/anshulkundaje/projects/blacklists) with [bedtools 2.19.1](http://bedtools.readthedocs.io/en/latest/),
* sorting and filtering of alignments with [samtools 1.6](http://samtools.sourceforge.net),
* cross-correlation computation with [spp](https://github.com/kundajelab/phantompeakqualtools),
* peak calling with [MACS 2.1.0](https://github.com/taoliu/MACS),
* irreproducible discovery rate computation with [idr](https://sites.google.com/site/anshulkundaje/projects/idr).

Briefly, in this workflow, reads are cleaned and trimmed using fastq-mcf from the ea-utils suite v1.1.2 to remove adapters, low quality bases and reads, and discard reads shorter than 25 bp after filtering. Reads are then aligned to the human reference genome (hg19) with bowtie v1.1.1using best matches parameters (bowtie -v 2 -m 1 --best --strata). Alignment files are further processed with samtools v1.2 and PicardTools v1.130 to flag PCR and optical duplicates and remove alignments located in Encode blacklisted regions. Fragment size is estimated *in silico* for each library using spp v1.10.1. Enriched regions are identified for each replicate independently with MACS v2.1.0 with non-IPed genomic DNA as a control (macs2 callpeak --nomodel --shiftsize --shift-control --gsize hs -p 1e-1). These relaxed peak lists are then processed through the irreproducible discovery rate (IDR) pipeline to generate an optimal and reproducible set of peaks for each histone modification type and each time point.

### Compare PIQ with ChIP-seq

```{r}
##--> Loading libraries
library(circlize)
library(limma)
library(plotrix)
library(gplots)
library(ComplexHeatmap)
library(rtracklayer)
library(fgsea)

compare_piq_chipseq <- function(tf, pwm, day) {
  
  ##--> Import ATAC-seq goldstandard peakset
  atacseq_gold <- import("./data/ATAC-seq/GOLDSTANDARD_ATACSEQ_ALL_DAY_IDR.bed")
  
  ##--> Import PIQ prediction
  root <- paste("./data/PIQ/RAW/", day, sep = "")
  piq_file <- list.files(path = root, pattern = "-calls.all.csv", full = TRUE)
  piq_file <- piq_file[grep(pwm, piq_file)]
  
  piq_call  <- do.call("rbind", lapply(piq_file, function(x) read.csv(x)))[,c(2,3,3,7)]
  colnames(piq_call) <- c("chr", "start", "end", "affinity")
  piq_call <- GRanges(piq_call)
  piq_call <- piq_call[-which(seqnames(piq_call) == "chrY" |seqnames(piq_call) == "chrX")]
  
  ##--> Select PWM matches in ATAC-seq peaks
  # and with an affinity score > 0.7 (bound sites)
  piq_call_1 <- subsetByOverlaps(piq_call, atacseq_gold)[which(subsetByOverlaps(piq_call, atacseq_gold)$affinity > .7)]
  
  ##--> Select randomnly 3 time more PWM matches 
  # with an affinity < 0.7 (unbound sites)
  piq_call_2 <- sample(piq_call[piq_call$affinity < .7], length(piq_call_1)*3)
  
  ##--> Fuse the two sets of PWM matches and format
  call <- unique(c(piq_call_1, piq_call_2)  + 1000)
  call <- call[rev(order(call$affinity))]
  
  ##--> Bin ChIP-seq peak 
  tile = 100
  call_split <- tile(call, n = tile)
  call_split <- unlist(call_split)
  
  ##--> Import normalized BigWig files for ChIP-seq and ATAC-seq
  bw_chipseq <- list.files("./data/ChIP-seq/",
                           pattern = paste(tf, ".bw", sep = ""),
                           full.names = TRUE)
  bw_atacseq <- list.files(path ="./data/ATAC-seq/",
                           pattern = paste(day, ".bw", sep = ""),
                           full = TRUE)
  bw <- c(bw_chipseq, bw_atacseq)
  
  ##--> Compute signals in bound and unbound PWM matches peaks
  counts <- matrix(NA, nrow = length(call_split), ncol = length(bw))
  
  for(i in c(1 : length(bw))) {
    print(i)
    coverage <- import(bw[i], as = 'RleList')
    
    for(chr in unique(seqnames(call_split))) {
      
      counts[which(seqnames(call_split) == chr) , i] <- mean(Views(coverage[[which(names(coverage) == chr)]],
                                                                   ranges(call_split[seqnames(call_split) == chr])))
      
    }
  }
  
  ##--> Transform into matrix
  count_list <- list()
  count_list[[1]] <- t(matrix(counts[,1], nrow = tile))
  count_list[[2]] <- t(matrix(counts[,2], nrow = tile))
  
  ##--> Ordering according to ChIP-seq signal around PWM match
  order  <- rev(order(rowSums(count_list[[1]][,c(45:55)])))
  order <- setdiff(order, which(rowSums(count_list[[1]] [,c(45:55)])==0))
  
  
  ##--> Defining the color scheme
  color_1 <- colorRamp2(seq(from = quantile(unlist(count_list[[1]]), probs = seq(0, 1, 0.05),na.rm = TRUE)[4],
                            to =   quantile(unlist(count_list[[1]]), probs = seq(0, 1, 0.05),na.rm = TRUE)[20],
                            length.out = 30),
                        smoothColors("#356FB2",28,"#EBE517"))
  color_2 <- colorRamp2(seq(from = quantile(unlist(count_list[[2]]), probs = seq(0, 1, 0.05),na.rm = TRUE)[4],
                            to =   quantile(unlist(count_list[[2]]), probs = seq(0, 1, 0.05),na.rm = TRUE)[20],
                            length.out = 30),
                        smoothColors("#356FB2",28,"#EBE517"))
  
  ##--> Defining annotation
  split <- data.frame(Affinity = call$affinity[order])
  split$State <- rep("UNBOUND", length(split))
  split$State[which(split$Affinity > .7)] <- "BOUND"
  split_1 <- split$State
  split_2 <- ComplexHeatmap::rowAnnotation(df = data.frame(Change = split_1), 
                                           col = list(Change = c("UNBOUND" =  "firebrick1", "BOUND" = "chartreuse3")),
                                           width = unit(.3, "cm"),
                                           show_legend = FALSE)
  
  ##--> Customized heatmap function
  heatmap_custom <- function(x, y) {
    
    Heatmap(x, cluster_rows = FALSE,
            cluster_columns = FALSE,
            gap = unit(0, "mm"),
            col = y,
            combined_name_fun = NULL,
            show_heatmap_legend = FALSE)
    
  } 
  
  ##--> Build the ComplexHeatmap objects
  heatmap_piq <- split_2 + 
    heatmap_custom(count_list[[1]][order,], color_1) +
    heatmap_custom(count_list[[2]][order,], color_2)
  heatmap_piq <- grid.grabExpr(draw(heatmap_piq))
  
  ##--> Set enrichment analysis
  
  # Compute summed ChIP-seq signal at PWM match
  chipseq_signal <- rowSums(count_list[[1]][,c(45:55)])[order]
  names(chipseq_signal) <- c(1:length(chipseq_signal))
  
  # Correlate "bound" status and ChIP-seq signal with SEA
  piq_bound <- split$State
  piq_bound <- list(BOUND = as.character(which(piq_bound == "BOUND")))
  fsea_bound <- fgsea(pathways = piq_bound,
                      stats = chipseq_signal,
                      nperm = 1000,
                      minSize = 1,
                      maxSize = Inf)
  fsea_bound_plot <- plotEnrichment(piq_bound$BOUND, chipseq_signal)
  
  # Correlate "unbound" status and ChIP-seq signal with SEA
  piq_unbound  <- split$State
  piq_unbound  <- list(UNBOUND = as.character(which(piq_unbound == "UNBOUND")))
  fsea_unbound <- fgsea(pathways = rev(piq_unbound ),
                        stats = rev(chipseq_signal),
                        nperm = 1000,
                        minSize = 1,
                        maxSize = Inf)
  fsea_unbound_plot <- plotEnrichment(piq_unbound$UNBOUND, chipseq_signal)
  
  # FSEA histogram
  fsea_combine_hist <- data.frame(Rank = c(1:length(split$State)) ,
                                  Type = as.numeric(as.factor(split$State)))
  fsea_combine_hist$Type <- factor(fsea_combine_hist$Type,
                                   labels = c("Bound", "Unbound"))
  
  fsea_hist <- ggplot(fsea_combine_hist, aes(x = Rank, group = Type)) + 
    geom_histogram(aes(group=Type, colour=Type, fill=Type),
                   alpha = 0.6, bins = 100, position = "fill") +
    theme_bw() + 
    scale_color_manual(values = c("firebrick1", "chartreuse3")) +
    scale_fill_manual(values = c("firebrick1", "chartreuse3")) +
    scale_x_reverse() + scale_y_reverse() + coord_flip() +
    theme(legend.position="none")
  
  # FSEA enrichment plot
  fsea_combine_plot <- rbind(fsea_bound_plot$data, fsea_unbound_plot$data)
  fsea_combine_plot$Type <- c(rep("BOUND", nrow(fsea_bound_plot$data)),
                              rep("UNBOUND", nrow(fsea_unbound_plot$data)))
  fsea_plot <- ggplot(fsea_combine_plot, aes(x=x, y=y, color = Type)) +
    geom_line() +
    theme_bw()  +
    scale_color_manual(values = c("firebrick1", "chartreuse3")) +
    coord_flip() + scale_x_reverse()  + theme(axis.title.y=element_blank()) +
    ylab("GSEA enrichment")
  
  return(list(heatmap_piq, fsea_plot, fsea_hist))

}

rela_comp <- compare_piq_chipseq("RELA", "RELA", "144H")
jun_comp <- compare_piq_chipseq("JUN", "JUNB", "144H")
fosl2_comp <- compare_piq_chipseq("FOSL2", "FOSL2", "144H")
```

```{r, fig.cap="Comparison between PIQ prediction and ChIP-seq for RELA, JUN and FOSL2"}
##--> RELA
grid.arrange(rela_comp[[3]], rela_comp[[1]], rela_comp[[2]],
             nrow = 1,
             top = textGrob("RELA - Comparison PIQ vs ChIP-seq"))

##--> JUN
grid.arrange(jun_comp[[3]], jun_comp[[1]], jun_comp[[2]],
             nrow = 1,
             top = textGrob("JUN - Comparison PIQ vs ChIP-seq"))

##--> FOSL2
grid.arrange(fosl2_comp[[3]], fosl2_comp[[1]], fosl2_comp[[2]],
             nrow = 1,
             top = textGrob("FOSL2 - Comparison PIQ vs ChIP-seq"))
```

# TF footprint characterization

## Computation of TF metrics

For each transcription factor, we computed the chromatin opening index, the motif dependence and the chromatin dependence as described in (Sherwood et al., 2014).

* The chromatin opening index is defined as the expected change in reads at bound occurrences with respect to baseline at $j^{th}$ binding site for the the transcription factor $m$ , in the experiment $k$, with $c_i^k$ the number of ATAC-seq read at the base $i$ in the experiment $k$, $p_j^k$ the PIQ estimated binding probability at site $j$, $x_j$ the coordinate of binding site $j$, and $S_m$ the set of binding sites for transcription factor $m$. It is defined as 

$$p_m= \frac{1}{399 k} \sum_{i=-199}^{199} \sum_{k} \left( \frac{\sum_{j\in S_m}c^{k+1}_{x_{j+i}}*p^k_i}{\sum_{j\in S_m}c^{k+1}_{x_{j+i}}*\left(1-p^k_i\right)} \right)$$ 

* The motif dependence of a factor $m$ is defined as the Pearsons correlation between the logistic binding probability and the PWM score $z$ at site $j$, defined as 

$$ b_m = cor(logit(p^k_j), z_j)$$

* The chromatin dependence is defined as the Pearsons correlation between logistic binding and the log number of ATAC-seq reads within a 399 bp window, computed as 

$$ c_m = cor(logit(p^k_j * log(\sum_{i=-199}^{199} c^{k}_{x_{j+i}})$$

**NB:** To compute these metrics, we used the R script described in the following chunk which is not evaluated here to avoid computationnal burden. To rerun the analysis, we advise to create a stand-alone R script containing the following code chunk and execute computations from the terminal using the `Rscript` command. This script take as an input the JASPAR PWM ID (*e.g.*: MA0911.1) and the time point reference (*e.g.*: D0).

```{r, eval = FALSE}
#--> Loading libraries
library(GenomicRanges)
library(GenomicAlignments)
library(genomation)
library(biovizBase)
library(BSgenome.Hsapiens.UCSC.hg19)
library(gtools)

#--> Input arguments
arg  <- commandArgs(TRUE)
arg1 <- arg[1] # e.g. MA0911.1
arg2 <- arg[2] # D0, D1, ..., D6
arg1 <- gsub("\\.", "", arg1)
print(Arg1)

#--> Import de JASPAR DB

importJaspar <- function(file=myloc) {
  vec <- readLines(file)
  vec <- gsub("\\[|\\]", "", vec)
  start <- grep(">", vec); end <- grep(">", vec) - 1
  pos <- data.frame(start=start, end=c(end[-1], length(vec)))
  pwm <- sapply(seq(along=pos[,1]), function(x) vec[pos[x,1]:pos[x,2]])
  pwm <- sapply(seq(along=pwm), function(x) gsub( "^ *|(?<= ) | *$", "", pwm[[x]], perl = TRUE))
  pwm <- sapply(seq(along=pwm), function(x) strsplit(pwm[[x]], " "))
  pwm <- sapply(seq(along=start), function(x) matrix(as.numeric(t(as.data.frame(pwm[(pos[x,1]+1):pos[x,2]]))[,-1]),nrow=4, dimnames=list(c("A", "C", "G", "T"), NULL)))
  names(pwm) <- gsub(">", "", vec[start])
  return(pwm)
}

pwm <- importJaspar(file="./tools/jaspar_eukaryotes_519_TF_PIQ_FORMATED.txt")

#--> Setups = PWM information
names <- names(pwm)
id <- gsub("\\.", "", gsub("_.*", "", names))  # format PWM names
id_target <- gsub("\\.", "", gsub("_.*", "", arg1))
id_pos <- grep(id_target , id)                 # get the position of the analyzed PWM on the JASPAR DB
id_length <- round(dim(pwm[[id_pos]])[2]/2)*2  # get the size of the PWM

#--> Setups = paths to data
root <- "/data/PIQ/ORIGINAL/"
root <- paste(root, arg2, "", sep = "")

#--> Import PIQ CSV files with coordinates and statistics for candidate sites
piq_call_files <- list.files(path = root, pattern = "-calls.all.csv", full = TRUE)
piq_call <- read.csv(piq_call_files[grep(id_target, piq_call_files)[1]], sep = ",")[,-1]

#--> Reformat target name
id_2 <- gsub("^.*_", "", names)
id_target <- id_2[match(id_target, id)]

#--> Convert the table to GRanges
piq_call_gr <- GRanges(seqnames = piq_call$chr, IRanges(start = piq_call$coord, end = piq_call$coord))
piq_call_gr$PWM <- piq_call$pwm
piq_call_gr$PIQ_purity <- piq_call$purity

#--> Resize PWM hits
piq_call_gr <- piq_call_gr + 199

#--> Identify the top 10000 hits based on FIMO score 
piq_call_gr_sel <- order(-piq_call_gr$PWM)[1:min(10000, length(piq_call_gr$PWM))]

#--> Split selected hits in 1bp windows
piq_call_gr_split <- tile(piq_call_gr, n = width(piq_call_gr)[1]) 
piq_call_gr_split <- flatGrl(piq_call_gr_split)

#--> Setup input bam file
bam_root <- c("./data/ATAC-seq/")
bam <- c(paste(bam_root,"RASOIS_ATACSEQ_T0_REP1.bam", sep = ""),
         paste(bam_root,"RASOIS_ATACSEQ_T0_REP2.bam", sep = ""),
         paste(bam_root,"RASOIS_ATACSEQ_T0_REP3.bam", sep = ""))
target_ranges <- piq_call_gr

#--> Import bam files (split FW and RV reads)
param_FW <- ScanBamParam(what = "pos",
                         which = piq_call_gr + 20000,
                         scanBamFlag(isMinusStrand = FALSE))
param_RV <- ScanBamParam(what = "pos",
                         ,
                         scanBamFlag(isMinusStrand = TRUE))

#--> Import alignments 
reads_FW <- GRanges(unlist(GAlignmentsList(bplapply(bam,
                                                    readGAlignments,
                                                    BPPARAM = MulticoreParam()))))
reads_RV <- GRanges(unlist(GAlignmentsList(bplapply(bam,
                                                    readGAlignments,
                                                    BPPARAM = MulticoreParam()))))

#--> Shift and resize alignements to focus on Tn5 cut sites
reads_FW <- GRanges(seqnames = seqnames(reads_FW),
                    IRanges(start = start(reads_FW),
                            end = end(reads_FW)),
                    strand = strand(reads_FW))
reads_RV <- GRanges(seqnames = seqnames(reads_RV),
                    IRanges(start = start(reads_RV),
                            end = end(reads_RV)),
                    strand = strand(reads_RV))
bases_FW <- shift(resize(reads_FW, 1), -4)
bases_RV <- shift(resize(reads_RV, 1), -4)

#--> Generate binding profiles for bound sites (FW)
count_test_FW <- summarizeOverlaps(piq_call_gr_split,
                                   bases_FW,
                                   ignore.strand = FALSE,
                                   inter.feature = FALSE)
count_test_FW <- assay(count_test_FW)
count_test_FW <- t(matrix(count_test_FW,
                          ncol = dim(count_test_FW)[1] / width(piq_call_gr)[1]))

#--> Generate binding profiles for bound sites (RV)
count_test_RV <- summarizeOverlaps(piq_call_gr_split,
                                   bases_RV,
                                   ignore.strand = FALSE,
                                   inter.feature = FALSE)
count_test_RV <- assay(count_test_RV)
count_test_RV <- t(matrix(count_test_RV,
                          ncol = dim(count_test_RV)[1] / width(piq_call_gr)[1]))

#--> Define region to compute local background for each PWM hit
background_gr <- promoters(piq_call_gr[piq_call_gr_sel],
                           upstream = 0,
                           downstream = 10000) - width(piq_call_gr[piq_call_gr_sel])
background_gr_split <- tile(background_gr, n = width(background_gr)[1])
background_gr_split <- flatGrl(background_gr_split)

#--> Define the binding profile in background regions (FW)
count_background_FW <- summarizeOverlaps(background_gr_split,
                                         bases_FW,
                                         ignore.strand = FALSE,
                                         inter.feature = FALSE)
count_background_FW <- assay(count_background_FW)
count_background_FW <- t(matrix(count_background_FW,
                               ncol = dim(count_background_FW)[1] / width(background_gr)[1]))

#--> Define the binding profile in background regions (FW)
count_background_RV <- summarizeOverlaps(background_gr_split,
                                         bases_RV,
                                         ignore.strand = FALSE,
                                         inter.feature = FALSE)
count_background_RV <- assay(count_background_RV)
count_background_RV <- t(matrix(count_background_RV,
                               ncol = dim(count_background_RV)[1] / width(background_gr)[1]))


#--> Compute the chromatin opening index
count_background_sum_FW <- rowSums(count_background_FW[,10000/2 + (-199:199)])
count_test_sum_FW <- rowSums(count_test_FW)
count_background_sum_RV <- rowSums(count_background_RV[,10000/2 + (-199:199)])
count_test_sum_RV <- rowSums(count_test_RV)

p_rank_piq <- 1 / (1+ exp(-log(piq_call_gr$PIQ_purity) + log(1 - piq_call_gr$PIQ_purity) +1 ))

coi_FW <- log(
              (sum(count_test_sum_FW[piq_call_gr_sel] * p_rank_piq[piq_call_gr_sel]) /  sum(p_rank_piq[piq_call_gr_sel])) / 
              (mean(count_background_sum_FW))) / log(2)

coi_RV <- log(
              (sum(count_test_sum_RV[piq_call_gr_sel] * p_rank_piq[piq_call_gr_sel]) /  sum(p_rank_piq[piq_call_gr_sel])) / 
              (mean(count_background_sum_RV))) / log(2)

PI_FW_PIQ <- log((sum(count_test_sum_FW[piq_call_gr_sel]*p_rank_piq[piq_call_gr_sel])/sum(p_rank_piq[piq_call_gr_sel]))/(mean(count_background_sum_FW)))/log(2)

coi <- (coi_FW + coi_RV)*2

#--> Compute the asymmetry index
ai <- coi^2 * abs(coi_FW - coi_RV )

#--> Compute the chromatin dependence
cd <- cor(logit(p_rank_piq[piq_call_gr_sel], min = -100, max = 100),
          (count_test_sum_FW[piq_call_gr_sel] + count_test_sum_RV[piq_call_gr_sel]))

#--> Compute the motif depedence
md <- cor(logit(p_rank_piq[piq_call_gr_sel], min = -100, max = 100),
          piq_call_gr$PWM[piq_call_gr_sel])

#--> Put all the results together and output
res <- data.frame(Chromatin_Opening_I = c(coi),
                  Asymmetry_I = c(zi),
                  Chrom_Dependence = c(cd),
                  Motif_Dependence = c(md),
                  row.names = c("PIQ"))
write.table(RES, paste("./data/PIQ/TF_METRICS/", paste(Arg2, ID_Target, sep = "_"), "_SI.txt", sep = ""), quote = F, row.names = T, col.names = T)
```

## Classification of TFs

```{r, fig.cap="Chromatin dependence versus chromatin opening index for all motifs"}
#--> Load libraries
library(plyr)

#--> List files with TF metrics for each PWM and each time point
file_comp <- list.files("./data/PIQ/TF_METRICS/", pattern = "_COI.txt", full = TRUE)

#--> Import data
data_comp <- do.call("rbind",
                     lapply(file_comp, function(fn) {
                            data.frame(Filename=fn,
                                       read.csv(fn,
                                                header = FALSE,
                                                sep= "",
                                                skip = 1))
                      }))

colnames(data_comp) <- c("Filename", "V1", "Chromatin_Opening_I",
                         "Asymmetry_I", "Chrom_Dependence", "Motif_Dependence")

#--> Format
name <- gsub("./data/PIQ/TF_METRICS/", "", data_comp[,1])
name <- gsub ("_COI.txt", "", name)
name <- gsub("D[012346]_", "", name)

day <-  gsub("./data/PIQ/TF_METRICS/", "", data_comp[,1])
day  <- gsub ("_.*_COI.txt", "", day)
data_comp$TF <- name
data_comp$Day <- day

#--> Compute the mean of each statistcs for each TF over time point
cd_mean <- ddply(data_comp, .(TF), summarize, mean=mean(Chrom_Dependence))
coi_mean <- ddply(data_comp, .(TF), summarize, mean=mean(Chromatin_Opening_I))

#--> Pepare data.frame
piq_profile <- data.frame(TF = unique(name), CD = cd_mean$mean, COI = coi_mean$mean)

#--> Apply empirical threshold to define migrants, settlers and pioneers
piq_profile$Type <- ""
piq_profile$Type[which(piq_profile$CD < 0.4)] <- "Migrants"
piq_profile$Type[which(piq_profile$CD >= 0.4 & piq_profile$COI < 4)] <- "Settlers"
piq_profile$Type[which(piq_profile$CD >= 0.4 & piq_profile$COI >= 4)] <- "Pioneers"
piq_profile <- unique(piq_profile)
piq_profile$Type <- factor(piq_profile$Type, levels = c("Pioneers", "Settlers", "Migrants"))

#--> Build the chromatin opening index vs chromatin dependance plot
ggplot(piq_profile, aes(x = CD, y = COI, color = Type, group = Type)) + 
  geom_point(size = 1) +
  xlab("PIQ motif dependence") +
  ylab("PIQ chromatin opening index")  +
  scale_colour_manual(values = c("#4682B4", "#FF3030", "#A2CD5A")) +
  geom_hline(yintercept = 4, linetype  = "dashed") +
  geom_vline(xintercept = 0.4, linetype  = "dashed") +
  theme_classic()
```

## Metaprofiles around TF binding sites

### Create a function to build Tn5 and nucleosome metaprofiles

```{r}

tf_metaprofiles <- function(pwm_id, day) {

  ##--> Format PWD ID
  pwm_id <- gsub("\\.", "", pwm_id)
  
  ##--> Build function to import JASPAR DB
  importJaspar <- function(file=myloc) {
    vec <- readLines(file)
    vec <- gsub("\\[|\\]", "", vec)
    start <- grep(">", vec); end <- grep(">", vec) - 1
    pos <- data.frame(start=start, end=c(end[-1], length(vec)))
    pwm <- sapply(seq(along=pos[,1]), function(x) vec[pos[x,1]:pos[x,2]])
    pwm <- sapply(seq(along=pwm), function(x) gsub( "^ *|(?<= ) | *$", "", pwm[[x]], perl = TRUE))
    pwm <- sapply(seq(along=pwm), function(x) strsplit(pwm[[x]], " "))
    pwm <- sapply(seq(along=start), function(x) matrix(as.numeric(t(as.data.frame(pwm[(pos[x,1]+1):pos[x,2]]))[,-1]),nrow=4, dimnames=list(c("A", "C", "G", "T"), NULL)))
    names(pwm) <- gsub(">", "", vec[start])
    return(pwm)
  }
  
  ##--> Import JASPAR DB
  pwm <- importJaspar(file="./tools/jaspar_eukaryotes_519_TF_PIQ_FORMATED.txt")
  names <- names(pwm)
  
  ##--> Get information on the pwm (TF names, motif length)
  id <- gsub("\\.", "", gsub("_.*", "", names))
  id_target <- pwm_id
  id_pos <- match(id_target, id)
  id_length <- round(dim(pwm[[id_pos]])[2]/2)*2
  
  ##--> Set up the path to access PIQ .bed files
  root <- "./data/PIQ/ORIGINAL/"
  bed_files <- list.files(path = root , pattern =   paste(".*_IDR_", "D0", ".*.bed", sep = ""), full = T)
  
  ##--> Process the candidate binding sites
  call  <- read.table(bed_files[grep(id_target, bed_files)[1]])
  call_gr <- GRanges(seqnames = as.character(call[,1]), 
                     IRanges(start =call[,2] - (500 + id_length/2 - 1),
                             end = call[,2] + (500 + id_length/2)))
  call_split <- tile(call_gr, n = width(call_gr)[1])
  call_split <- flatGrl(call_split)
  
  ##--> Convert PWM ID to HGNC
  id_hgnc <- gsub("^.*_", "", names)
  id_hgnc_tf <- id_hgnc[match(id_target, id)]
  
  ##--> List .bam files
  bam_root <- c("./data/ATAC-seq/")
  bam <- c(paste(bam_root,"RASOIS_ATACSEQ_",day,"_REP1.bam", sep = ""),
           paste(bam_root,"RASOIS_ATACSEQ_",day,"_REP2.bam", sep = ""),
           paste(bam_root,"RASOIS_ATACSEQ_",day,"_REP3.bam", sep = ""))
  
  param_FW <- ScanBamParam(what = "pos",
                           scanBamFlag(isMinusStrand = F),
                           which = call_gr + 1000)
  param_RV <- ScanBamParam(what = "pos", scanBamFlag(isMinusStrand = T),
                           which = call_gr + 1000)
  
  ##--> Prepare environment for storing aligments
  count_FW_all_piq <- list()
  count_RV_all_piq <- list()
  count_FW_all_plot <- list()
  count_RV_all_plot <- list()
  
  bases_FW_all <- list()
  bases_RV_all <- list()
  
  ##--> Import alignments 
  reads_FW <- GRanges(unlist(GAlignmentsList(bplapply(bam,
                                                      readGAlignments,
                                                      BPPARAM = MulticoreParam()))))
  reads_RV <- GRanges(unlist(GAlignmentsList(bplapply(bam,
                                                      readGAlignments,
                                                      BPPARAM = MulticoreParam()))))
  
  reads_FW <- GRanges(seqnames = seqnames(reads_FW),
                      IRanges(start = start(reads_FW),
                              end = end(reads_FW)),
                      strand = strand(reads_FW))
  
  reads_RV <- GRanges(seqnames = seqnames(reads_RV),
                      IRanges(start = start(reads_RV),
                              end = end(reads_RV)),
                      strand = strand(reads_RV))
  
  ##--> Shift and resize to focuse on the 5' end (Tn5 cuts)
  bases_FW <- GenomicRanges::shift(GenomicRanges::resize(reads_FW, 1),-4)
  bases_FW_all <- bases_FW
  bases_RV <- GenomicRanges::shift(GenomicRanges::resize(reads_RV, 1),-4)
  bases_RV_All <- bases_RV
  
  ##--> Select bound PWM occurences
  bound_all_RV <- list()
  bound_all_FW <- list()
  n_bound <- list()
  
  count_bound_FW <- summarizeOverlaps(call_split, bases_FW_all, yieldSize = 100000, ignore.strand = F, inter.feature = FALSE)
  count_bound_FW <- matrix(assay(count_bound_FW), ncol = width(call_gr)[1], byrow = TRUE)
  count_bound_RV <- summarizeOverlaps(call_split, bases_RV_All, yieldSize = 100000, ignore.strand = F, inter.feature = FALSE)
  count_bound_RV <- matrix(assay(count_bound_RV), ncol = width(call_gr)[1], byrow = TRUE)
  
  ##--> Define a quantile threshold
  thresh_1 <- max(quantile(cbind(count_bound_RV, count_bound_FW), 0.999), 1L)
  cuts_1 <- pmin(cbind(count_bound_RV, count_bound_FW), thresh_1 )
  
  ##--> Subset and format the data
  count_bound_FW <- cuts_1[,c(1:(ncol(cuts_1)/2))]
  count_bound_RV <- cuts_1[,c(((ncol(cuts_1)/2+1) : ncol(cuts_1)))]
  count_bound_FW_2  <- count_bound_FW[,c((1):(500 + id_length))]
  count_bound_RV_2  <- count_bound_RV[,c(501:(1000 + id_length))]
  
  ##--> Compute aveage Tn5 cut profiles
  mean_bound_FW <- apply(count_bound_FW_2 , 2, mean)
  mean_bound_RV <- apply(count_bound_RV_2, 2, mean)
  
  ##--> Formating for ggplot
  mean_cut <- data.frame(Count = c(mean_bound_FW, mean_bound_RV),
                         Distance = rep(c(seq(-500 - id_length/2 + 1, id_length/2),
                                          seq(-id_length/2 + 1 ,500 + id_length/2))),
                         Type = c(rep("Bound", length(c(unlist(mean_bound_FW))) +
                                        length(c(unlist((mean_bound_RV)))))),
                         Strand = c(rep("FW", length(unlist(mean_bound_FW))),
                                    rep("RV", length(unlist(mean_bound_RV)))))
  
  labels <-  c(paste(day, " (N=", length(call_gr), ")", sep = ""))
  
  ##--> Tn5 cut profile
  plot_cut <- ggplot(mean_cut, aes(x = Distance, y = Count, group = Strand, color = Strand)) +
    geom_line(aes(colour= Strand), size = .3) + 
    theme_bw() +
    scale_color_manual(values=c("grey60", "grey10")) +
    theme(legend.position="none") +
    ggtitle(paste(id_target, "Average Tn5 insertion profil \n (Bound vs Unbound according to PIQ)", sep = " - "))
  
  
  ##--> Import nucleosome density track
  bw_nuc <- paste(bam_root,"RASOIS_ATACSEQ_",day,"_REP123_NUCLEO_OCC.bw", sep = "")
  
  ##--> Compute nucleosome occupancy around bound instances 
  count_nuc <- matrix(NA, nrow = length(call_split), ncol = 1)
  chrs <- c(paste("chr", seq(1, 22), sep = ""))
  
  coverage <- import(bw_nuc, as = 'RleList')
  
  for(chr in chrs) {
    count_nuc[which(seqnames(call_split) == chr) , 1] <-
      mean(Views(coverage[[which(names(coverage) == chr)]], ranges(call_split[seqnames(call_split) == chr])))
  }
  
  ##--> Format nucleosome occupancy matrix
  count_nuc_2  <- t(matrix(count_nuc[,1], nrow = length(call_split) / length(call_gr)))
  count_nuc_mean <- colMeans(count_nuc_2, na.rm = TRUE)
  
  ##--> Formating for ggplot
  count_nuc_mean <- data.frame(Nuc = c(unlist(count_nuc_mean)), 
                               Distance = ((-500 - id_length/2  ) : (500 + id_length/2 - 1)))
  
  ##--> Ncleosome density profile
  plot_nuc <- ggplot(count_nuc_mean, aes(x = Distance, y = Nuc)) +
    geom_smooth(color = "orange", size = 2, level = .99) +
    theme_bw()
  
  metaplot <-  plotflow::ggdual_axis(plot_cut,plot_nuc)
  return(metaplot)
}

meta_FOSL1 <- tf_metaprofiles("MA04771", "T0")
meta_RELA <- tf_metaprofiles("MA01071", "T0")
meta_SREBF1 <- tf_metaprofiles("MA05951", "T0")
```

### Build metaprofiles

```{r, fig.cap="ATAC-seq (grey and black lines) and nucleosome (red line) profiles at the FOSL1, RELA and SREBF1 bound sites", fig.height=12}

grid.arrange(meta_FOSL1, meta_RELA, meta_SREBF1, ncol = 1)

```


## Exploration of the TF landscape

Using the package `r Biocpkg("annotatr")` we first created an annotation datasets combining coordinates for *hg19* promoters, 3UTRs, exons, introns and intergenic regions as defined in UCSC, as well as our custom set of enhancers (intersection of the global sets of reproducible H3K4me1 peaks with global sets of reproducible H3K27ac peaks, to focus on enhancers that are active at least in one time point during RAS OIS in WI38 fibroblasts). We then annotated the PIQ binding occurrences for each of the PWM independently, and for the 6 time-points independently. Data were further normalized, to account for differences between time-points in the total number of bound occurrences summed across PWM, and finally they were converted to frequencies. We filtered TFs for which less than 100 binding sites were identified throughout the entire timecourse. TFs were ordered according to the proportions of binding sites located in promoters, introns or exons, and we finally computed the density in migrant, settler and pioneer factors along the ranking.

### Build the annotation set

```{r}
##--> Load libraries
library(GenomicRanges)
library(annotatr)
library(plyr)
library(reshape)
library(ggplot2)
library(rtracklayer)

##--> Create a function to import JASPAR DB
importJaspar <- function(file=myloc) {
  vec <- readLines(file)
  vec <- gsub("\\[|\\]", "", vec)
  start <- grep(">", vec); end <- grep(">", vec) - 1
  pos <- data.frame(start=start, end=c(end[-1], length(vec)))
  pwm <- sapply(seq(along=pos[,1]), function(x) vec[pos[x,1]:pos[x,2]])
  pwm <- sapply(seq(along=pwm), function(x) gsub( "^ *|(?<= ) | *$", "", pwm[[x]], perl = TRUE))
  pwm <- sapply(seq(along=pwm), function(x) strsplit(pwm[[x]], " "))
  pwm <- sapply(seq(along=start), function(x) matrix(as.numeric(t(as.data.frame(pwm[(pos[x,1]+1):pos[x,2]]))[,-1]),nrow=4, dimnames=list(c("A", "C", "G", "T"), NULL)))
  names(pwm) <- gsub(">", "", vec[start])
  return(pwm)
}

##--> Import JASPAR PWM and list TFs
pwm <- importJaspar(file="./tools/jaspar_eukaryotes_519_TF_PIQ_FORMATED.txt")
names <- names(pwm)

##--> Set path to ChIP-seq and PIQ data
root_chip <- "./data/ChIP-seq/"
root_piq <- "./data/PIQ/ORIGINAL/"

##--> Define K4Me1 positive regions at any given time point
chip_H3K4me1 <- list.files(path = root_chip,
                           full.names = TRUE,
                           pattern = "H3K4me1_MACS2_OPTIMAL_IDR.bed")
chip_H3K4me1 <- lapply(chip_H3K4me1, read.table)
chip_H3K4me1 <- GRangesList(lapply(chip_H3K4me1, function(x) {
                                     GRanges(seqnames = x$V1, IRanges(x$V2, x$V3))}))
chip_H3K4me1 <- lapply(chip_H3K4me1, GenomicRanges::reduce)

##--> Define K27Ac positive regions at any given time point
chip_H3K27ac <- list.files(path = root_chip,
                           full.names = TRUE,
                           pattern = "H3K27ac_MACS2_OPTIMAL_IDR.bed")
chip_H3K27ac <- lapply(chip_H3K27ac, read.table)
chip_H3K27ac <- GRangesList(lapply(chip_H3K27ac, function(x) {
                                     GRanges(seqnames = x$V1, IRanges(x$V2, x$V3))}))
chip_H3K27ac <- lapply(chip_H3K27ac, GenomicRanges::reduce)

##--> Define set of active enhancers
# as the overlap between H3K4me1 and H3K27ac at a given time point
active_enhancers <- reduce(unlist(GRangesList(GenomicRanges::intersect(chip_H3K4me1[[1]], chip_H3K27ac[[1]]),
                                              GenomicRanges::intersect(chip_H3K4me1[[2]], chip_H3K27ac[[2]]),
                                              GenomicRanges::intersect(chip_H3K4me1[[3]], chip_H3K27ac[[3]]))))
##--> Resize enhancers
active_enhancers  <- active_enhancers  + 1000
export(active_enhancers, "./data/ChIP-seq/custome_active_enhancers.bed", format = "BED")
read_annotations(con = "./data/ChIP-seq/custome_active_enhancers.bed", name = 'enhancers', genome = 'hg19')

##--> Build the annotation set
piq_annotation_set <- build_annotations(genome = "hg19",
                                    annotations = c("hg19_custom_enhancers", "hg19_genes_promoters",
                                                    "hg19_genes_3UTRs", "hg19_genes_exons",
                                                    "hg19_genes_introns", "hg19_genes_intergenic"))

```

### Annotate TF binding sites

```{r, fig.cap="The RAS OIS TF landscape in space and time"}
##--> Enrichment analysis over features for each time point and each TF

piq_annotation_raw <- list()
i <- 0

# Loop over time-points
for (day in c("D0", "D1", "D2", "D3", "D4", "D6")) {
  
  print(day)
  i <- i + 1
  
  # List PIQ files corresponding to each time point
  bed_files <- list.files(root_piq, full = TRUE, pattern = paste(".*_IDR_", day, ".*.bed", sep = ""))
  count_feature <- data.frame(matrix(ncol = 7))
  
  # Loop over TFs
  for (tf in names) {
    
    print(tf)
    id <- gsub("\\.", "", tf)
    id <- gsub("_.*", "", id)
    
    # Process the candidate binding sites
    bed_file_tf <- bed_files[grep(id, bed_files)]
    call_tf  <- adply(bed_file_tf, 1, read.table)
    call_tf_gr <- GRanges(seqnames = call_tf [,2],
                          IRanges(start = call_tf [,3], end = call_tf[,3]))
    length_call_tf <- length(call_tf_gr)
    
    # Set annotations to zero if the TF has less than 10 binding sites ...
    if (length_call_tf < 10) { Count_feature <- rbind(Count_feature, rep(0,6)) }
    
    # Else, go through the annotation process
    if (length_call_tf >= 10) {
      
      call_tf_annot <- annotate_regions(regions = call_tf_gr,
                                        annotations = piq_annotation_set,
                                        ignore.strand = TRUE,
                                        quiet = FALSE)
      call_tf_annot_table <- data.frame(summarize_annotations(call_tf_annot))[,2]
      count_feature <- rbind(count_feature,
                             c(length_call_tf, call_tf_annot_table))
      
    }
  }
  
  piq_annotation_raw[[i]] <- count_feature
}
```

### Generate the TF landscape map

```{r, fig.cap="The RAS OIS TF landscape in space and time"}
##--> Load packages
library(gtable)
library(grid)
library(data.table)
library(scales)

##--> Format the output 
piq_annotation_raw_f <- do.call("rbind",  piq_annotation_raw)
piq_annotation_raw_f <- piq_annotation_raw_f[-which(is.na(piq_annotation_raw_f)),]
piq_annotation_raw_f$TF <- rep(names, 6)
piq_annotation_raw_f$Day <- rep(c("D0", "D1", "D2", "D3", "D4", "D6"), each = length(names))
colnames(piq_annotation_raw_f) <- c( "all", "enhancers","3UTRs",
                                     "exons","intergenic", "introns",
                                     "promoters", "TF", "Day")
piq_annotation_raw_f <-piq_annotation_raw_f[,c(8,9,1,5,2,6,4,3,7)]

##--> Remove redundant TFs
tf_redundant <- c("MA0807.1_TBX5", "MA0806.1_TBX4", "MA0805.1_TBX1",
                    "MA0849.1_FOXO6", "MA0848.1_FOXO4", "MA0847.1_FOXD2",
                    "MA0042.2_FOXI1", "MA0033.2_FOXL1")
piq_annotation_raw_f <- piq_annotation_raw_f[!piq_annotation_raw_f$TF %in% tf_redundant,]

##--> Compute %
piq_annotation_freq <- piq_annotation_raw_f
piq_annotation_freq$TF <- gsub("MA.*\\..*_", "", piq_annotation_freq$TF)
piq_annotation_freq[,c(4:9)] <- piq_annotation_freq[,c(4:9)] / apply(piq_annotation_freq[,c(4:9)], 1, sum) * 100

##--> Scale total count
sum <- c(unlist(by(piq_annotation_freq, piq_annotation_freq$Day, function(x) sum(x$all))))
piq_annotation_freq$sum <- sum[match(piq_annotation_freq$Day, names(sum))]
piq_annotation_freq$all <- round(piq_annotation_freq$all / piq_annotation_freq$sum * min(sum))

##--> Remove TF with few binding sites (< 100)
tf_remove <- unique(piq_annotation_freq[which(piq_annotation_freq$all < 300),1])
piq_annotation_freq <- piq_annotation_freq[!piq_annotation_freq$TF %in% tf_remove,]

##--> Reshape
piq_annotation_freq_melt <- melt(piq_annotation_freq[,-10])
colnames(piq_annotation_freq_melt) <- c("TF", "Day", "Features", "Count")
tf_count <- piq_annotation_freq_melt[which(piq_annotation_freq_melt$Feature == "all"),]
piq_annotation_freq_melt <- piq_annotation_freq_melt[-which(piq_annotation_freq_melt$Feature == "all"),]

##--> Define the TF ordering
tf_order <- piq_annotation_freq[which(piq_annotation_freq$Day == "D0"),]
tf_order <- tf_order[order(rowSums(tf_order[,c(7:9)]),decreasing = FALSE),]
piq_annotation_freq_melt$TF <- factor(piq_annotation_freq_melt$TF, levels = tf_order$TF)
tf_count$TF <- factor(tf_count$TF, levels = tf_order$TF)

##--> Define the Feature ordering
piq_annotation_freq_melt$Features <- factor(piq_annotation_freq_melt$Features,
                                            levels = rev(c("promoters", "3UTRs", "exons", 
                                                           "introns", "enhancers", "intergenic")))

##--> Generate side annotation with density in pionners / settlers / migrants
tf_classif <- read.table("./data/PIQ/ORIGINAL/PIQ_TF_CLASSIFICATION.txt", header =T)
tf_order_classif <- data.frame(TF = tf_order$TF)
tf_order_classif$Classif <- "black"
tf_order_classif$Classif[tf_order_classif$TF %in% tf_classif[which(tf_classif$Type == "Pioneers"),]$TF] <- "blue"
tf_order_classif$Classif[tf_order_classif$TF %in% tf_classif[which(tf_classif$Type == "Settlers"),]$TF] <- "red"
tf_order_classif$Classif[tf_order_classif$TF %in% tf_classif[which(tf_classif$Type == "Migrants"),]$TF] <- "green"
tf_order_classif$Type <- factor(tf_order_classif$Classif, levels = c("blue", "red", "green"))
tf_order_classif$Rank <- c(1:nrow(tf_order_classif))
tf_order_classif$Day <- "All"

##--> Density Plot
A <- ggplot(tf_order_classif, aes(x = Rank, group = Type)) + 
  geom_density(aes(group=Type, colour=Type, fill=Type), alpha = 0.6) +
  scale_fill_manual(values = c("#4682B4", "#FF3030", "#A2CD5A")) +
  scale_colour_manual(values = c("#4682B4", "#FF3030", "#A2CD5A")) +
  coord_flip() +
  facet_wrap(~Day) +
  scale_x_continuous(expand = c(0.0,0.0)) +
  scale_y_continuous(labels = scientific) +
  theme(panel.background = element_rect(fill = NA),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(fill = NA, colour = "grey50"),
        axis.text.x = element_text(angle = 90, hjust = 1),
        axis.title.x =  element_blank(),
        legend.position="none",
        strip.background = element_blank(),
        strip.text.x = element_blank())


##--> Annotation plot
B <- ggplot() + 
  geom_bar(data = piq_annotation_freq_melt, aes(y = Count, x = TF, col = Features, fill = Features), stat = "identity", position = "stack") +
  scale_color_manual(values = rev(c("brown3", "chocolate1", "whitesmoke", "gold", "darkolivegreen3", "steelblue"))) +
  scale_fill_manual(values = rev(c("brown3", "chocolate1", "whitesmoke", "gold", "darkolivegreen3", "steelblue"))) +
  facet_wrap(~Day, nrow = 1) +
  coord_flip() +
  theme(axis.text.y = element_text(colour=tf_order_classif$Classif)) +
  theme(panel.background = element_rect(fill = NA),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(fill = NA, colour = "grey50"),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x = element_text(angle = 90, hjust = 1),
        strip.background = element_blank(),
        strip.text.x = element_blank())

##--> Plot showing the number of binding siter per TF per day
C <- ggplot() + 
  geom_point(data = tf_count, aes(y = Count, x = TF), color = "black", size = .5) +
  facet_wrap(~Day, nrow = 1) +
  theme(panel.background = element_rect(fill = NA),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(fill = NA, colour = "grey50"),
        strip.background = element_blank(),
        strip.text.x = element_blank()) +
        coord_flip()

##--> Overlapping plots B and C
# To use facets with a dual y-axis ggplot, we used the tricks described at 
# https://stackoverflow.com/questions/26917689/how-to-use-facets-with-a-dual-y-axis-ggplot/40746716

p1 <- B
p2 <- C
g1 <- ggplotGrob(B)
g2 <- ggplotGrob(C)

# Get the locations of the plot panels in g1.
pp <- c(subset(g1$layout, grepl("panel", g1$layout$name), se = t:r))

# Overlap panels for second plot on those of the first plot
g <- gtable_add_grob(g1, g2$grobs[grepl("panel", g1$layout$name)], 
                     pp$t, pp$l, pp$b, pp$l)

# ggplot contains many labels that are themselves complex grob; 
# usually a text grob surrounded by margins.
# When moving the grobs from, say, the left to the right of a plot,
# Make sure the margins and the justifications are swapped around.
# The function below does the swapping.
# Taken from the cowplot package:
# https://github.com/wilkelab/cowplot/blob/master/R/switch_axis.R 

hinvert_title_grob <- function(grob){
  
  # Swap the widths
  widths <- grob$widths
  grob$widths[1] <- widths[3]
  grob$widths[3] <- widths[1]
  grob$vp[[1]]$layout$widths[1] <- widths[3]
  grob$vp[[1]]$layout$widths[3] <- widths[1]
  
  # Fix the justification
  grob$children[[1]]$hjust <- 1 - grob$children[[1]]$hjust 
  grob$children[[1]]$vjust <- 1 - grob$children[[1]]$vjust 
  grob$children[[1]]$x <- unit(1, "npc") - grob$children[[1]]$x
  grob
}

# Get the y axis title from g2
index <- which(g2$layout$name == "ylab-l") # Which grob contains the y axis title?   EDIT HERE
ylab <- g2$grobs[[index]]                  # Extract that grob
ylab <- hinvert_title_grob(ylab)           # Swap margins and fix justifications

# Put the transformed label on the right side of g1
g <- gtable_add_cols(g, g2$widths[g2$layout[index, ]$l], max(pp$r))
g <- gtable_add_grob(g, ylab, max(pp$t), max(pp$r) + 1, max(pp$b), max(pp$r) + 1, clip = "off", name = "ylab-r")

# Get the y axis from g2 (axis line, tick marks, and tick mark labels)
index <- which(g2$layout$name == "axis-l-1-1")  # Which grob
yaxis <- g2$grobs[[index]]                      # Extract the grob

# yaxis is a complex of grobs containing the axis line, the tick marks, and the tick mark labels.
# The relevant grobs are contained in axis$children:
#   axis$children[[1]] contains the axis line;
#   axis$children[[2]] contains the tick marks and tick mark labels.

# First, move the axis line to the left
# But not needed here
# yaxis$children[[1]]$x <- unit.c(unit(0, "npc"), unit(0, "npc"))

# Second, swap tick marks and tick mark labels
ticks <- yaxis$children[[2]]
ticks$widths <- rev(ticks$widths)
ticks$grobs <- rev(ticks$grobs)

# Third, move the tick marks
# Tick mark lengths can change. 
# A function to get the original tick mark length
# Taken from the cowplot package:
# https://github.com/wilkelab/cowplot/blob/master/R/switch_axis.R 
plot_theme <- function(p) {
  plyr::defaults(p$theme, theme_get())
}

tml <- plot_theme(p1)$axis.ticks.length   # Tick mark length
ticks$grobs[[1]]$x <- ticks$grobs[[1]]$x - unit(1, "npc") + tml

# Fourth, swap margins and fix justifications for the tick mark labels
ticks$grobs[[2]] <- hinvert_title_grob(ticks$grobs[[2]])

# Fifth, put ticks back into yaxis
yaxis$children[[2]] <- ticks

# Put the transformed yaxis on the right side of g1
g <- gtable_add_cols(g, g2$widths[g2$layout[index, ]$l], max(pp$r))
g <- gtable_add_grob(g, yaxis, max(pp$t), max(pp$r) + 1, max(pp$b), max(pp$r) + 1, 
                     clip = "off", name = "axis-r")

##--> Generate the plot
grid.arrange(A, g,  layout_matrix = rbind(c(1,rep(2,9)),c(1,rep(2,9))))
```

# Session Info
```{r, eval = TRUE, results = 'markup', message = TRUE, echo = TRUE,}
sessionInfo()
```
